---
title: "Exponential Smoothing Methods"
author: "Piotr Wieczorek"
date: "2023-06-27"
output: 
  html_document: 
    toc: yes
    toc_depth: 2
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message = FALSE)
```

```{r}
library(tidyverse)
library(forecast)
library(tseries)
library(expsmooth)
library(ggpubr)
library(png)
library(scales)
```


## **Simple Exponential Smoothing (SES)**

### This method is practical when there is no trend and no seasonality in the time series data. It essentialy gives weight (alpha) to past observations, where the "newest" observations get higher weights than the "oldest" data points in contradiction to the mean method, where all observatinos get the same weight. In SES the weights decrease exponentially. 

### Formula:

```{r,fig.width=5.5,fig.height=1.25}
ggplot() + 
    background_image(readPNG("images/ses_formula3.png"))
```

### The same formula  can be written as:

```{r,fig.width=6.5,fig.height=1.25}
ggplot() + 
    background_image(readPNG("images/ses_formula4.png"))
```



### We will use 'ausair' time series from the fpp2 package for the SES model

```{r}
ses_ts <- fpp2::ausair
```

```{r}
time(ses_ts)
```


```{r}
ses_ts <- window(ses_ts,start = 1989, end = 2016, frequency = 1)
ses_ts <- diff(ses_ts)
```


```{r}
autoplot(ses_ts)
```


```{r}
ses(ses_ts,h=5)
```


```{r}
autoplot(ses(ses_ts,h=5,alpha=0.2)) + 
  scale_x_continuous(breaks=seq(1990,2021,2))
```



## **Holt's method**

### For Holt's method, we will use 'unsetelec' time series from the expsmooth package.
### The data exhibits clear trend and no seasonal pattern, therefore Holt's method is suitable in this example.

### Formulas:
<h4>*Lt = level at time t</h4>
<h4>*Tt = trend at time t</h4>
<h4>*alpha and beta = smoothing parameters</h4>
<h4>*h = number of prediction steps ahead</h4>

```{r,fig.width=4,fig.height=2}
ggplot() + 
    background_image(readPNG("images/holts.png"))
```


```{r}
holt_ts <- expsmooth::usgdp
```

### The data ranges from Q1 1947 to Q1 2006. Given that it's financial data, values from half century ago most likely have 0 influence on values from 2000 + years. In fact, when holt's model was applied to the whole data set (meaning 1947-2006), the alpha value was equal to 0.9999, which indicates very strong influence of the most actual values and almost 0 influence of the oldest values. To avoid such high values of alpha and making the model more applicable, it was decided to take data points only from 1990-2006.

```{r}
time(holt_ts)
```


```{r}
holt_ts <- window(holt_ts,start = 1990, end = 2006, frequency = 4)
```


```{r}
autoplot(holt_ts)+
  scale_x_continuous(breaks=seq(1990,2006,1))
```



```{r,fig.width=10,fig.height=6}
autoplot(holt(holt_ts,h=12,damped = FALSE)) + 
  scale_y_continuous(labels = comma_format())+
  scale_x_continuous(breaks=seq(1990,2010,1))+
  theme_bw()
```


```{r}
holt_model <- holt(holt_ts,h=12,damped = FALSE)
```

```{r}
holt_model
```

```{r}
holt_model[['model']]
```

## **Holt-Winters' method**

### Holt-Winters' model is applicable if there is both trend and seasonality in the time series.

### Formulas for additive and multiplicative models:

<h4>*alpha is a smoothing parameter for the level component</h4>
<h4>*beta is a smoothing paramater for the trend component (the equation is the same as in the Holt's model)</h4>
<h4>*gamma is a smoothing paramater for the seasonal component</h4>


```{r,fig.width=4,fig.height=2}
ggplot() + 
    background_image(readPNG("images/holtwinters_formula.png"))
```



```{r,fig.width=4,fig.height=2}
ggplot() + 
    background_image(readPNG("images/holtwinters_multiplicative.png"))
```


```{r}
holt_winter <- fpp2::a10
```



```{r}
time(holt_winter)
```

```{r}
holt_winter <- window(holt_winter,start=2000,end=c(2008,6),frequency=12)
```

```{r}
holt_winter
```

### Data exhibits both trend and multiplicative seasonality

```{r}
autoplot(holt_winter)
```

```{r}
holt_winter_mod <- hw(holt_winter,seasonal = "multiplicative",h = 12)
```

```{r}
holt_winter_mod
```
```{r}
autoplot(holt_winter_mod) + 
  scale_x_continuous(breaks = seq(2000,2009,by=1))
```

```{r}
holt_winter_mod[['model']]
```


### References:
1. https://www.geeksforgeeks.org/exponential-smoothing-in-r-programming/
2. https://otexts.com/fpp2/holt-winters.html
3. https://otexts.com/fpp2/holt.html















