---
title: "Using sf and ggplot to create map plots in R"
author: "Piotr Wieczorek"
date: "2023-12-01"
output: 
  html_document: 
    toc: yes
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


```{r}
rm(list=ls())
```



## Libraries


```{r}
library(tidyverse)
library(scales)
library(ggmap)
library(mapproj)
library(maps)
library(mapproj)
library(readxl)
library(sf)
library(RColorBrewer)
```


## Reading the goegraphical data about the entire world

```{r}
world_tbl <- map_data("world") %>%
  as_tibble()
```


```{r}
head(world_tbl)
```


## Reading and cleaning the data - inflation rate in Europe in October, 2023


```{r}
inflation_october_2023 <- read_excel("inflacja_pazdziernik_2023.xlsx", 
    sheet = "Arkusz1")
```


```{r}
colnames(inflation_october_2023) <- c("region","inflation_rate_10_2023")
```


```{r}
inflation_october_2023 <- inflation_october_2023 %>%
  mutate(region = case_when(region == "Kosovo*" ~ "Kosovo",
                            region == "United Kingdom" ~ "UK",
                            TRUE ~ region))
```

```{r}
inflation_october_2023 <- world_tbl %>%
  left_join(inflation_october_2023, by = "region")
```


```{r}
inflation_october_2023 <- inflation_october_2023[complete.cases(inflation_october_2023[,
                                      c("inflation_rate_10_2023")]),]
```



### Map plot picturing inflation rate in Europe

```{r}
inflation_october_2023 <- inflation_october_2023 %>%
  filter(!(subregion == "Svalbard" & !(is.na(subregion))))
```




```{r,fig.width=12,fig.height=8}
inflation_october_2023 %>%
  ggplot(aes(x = long, 
             y = lat,
             group=group,
             text = paste0(region,
                          ": ",
                          inflation_rate_10_2023,"%"))) + 
  geom_polygon(aes(fill = inflation_rate_10_2023),color="black") +
  geom_path(color = "white", size = 0.5) +
  scale_fill_gradient(name = "inflation rate (%)", low = "green", high = "red", na.value = "gray50") + 
  labs(title = "Inflation Rate in Europe (annual rate of change) in October 2023",
       x = "",
       y= "") + 
  theme(rect  = element_blank(),
        plot.title = element_text(size=18),
        legend.title = element_text(size=16),
        legend.text = element_text(size=14)) 


```



## Reading the data about polish voivodeships coordinates


```{r}
voivodships <- st_read("poland_woj.json")
```

```{r}
voivodships$name
```

## Reading the data about unemployment rate in polish voivodeships

```{r}
unemp_10_2023 <- read_excel("stopa_bezrobocia_10_2023.xlsx", 
    sheet = "Sheet1", col_names = FALSE)
```

### Cleaning and joining the data

```{r}
colnames(unemp_10_2023) <- c("name","unemp_rate")
```

```{r}
unemp_10_2023 <- voivodships %>%
  left_join(unemp_10_2023, by = "name")
```

```{r}
class(unemp_10_2023)
```

```{r}
unemp_10_2023
```


### Map plot picturing unemployment rate within polish territory administrative division

```{r,fig.width=12,fig.height=8}
ggplot(unemp_10_2023) +
  geom_sf(aes(fill = unemp_rate)) +
  scale_fill_gradient(name = "Unemployment Rate (%)", low = "#CDBABA", high = "#BC353A") +
  labs(title = "Unemployment Rate (%) in Polish Voivodeships") +
  theme_minimal() + 
  theme(plot.title = element_text(size=18),
        legend.title = element_text(size=16),
        legend.text = element_text(size=14))
```


## Reading data about german bundeslands coordinates

```{r}
bundeslands <- st_read("germany-states.geojson")
```

```{r}
bundeslands$NAME_1
```


## Reading data about median of brutto salary in Germany and joining the data


```{r}
einkommen_median <- read_excel("bruttogehalt_median.xlsx", 
    sheet = "Sheet2", col_names = TRUE)
```

```{r}
einkommen_median
```

```{r}
einkommen_median <- bundeslands %>%
  left_join(einkommen_median, by = "NAME_1")
```

```{r}
einkommen_median[,c("NAME_1","BRUTTOMONATSEINKOMMEN_MEDIAN")]
```



### Map plot picturing median salary within german administrative territory division units


```{r,fig.width=12,fig.height=8}
ggplot(einkommen_median) +
  geom_sf(aes(fill = BRUTTOMONATSEINKOMMEN_MEDIAN)) +
  scale_fill_gradient(name ="mittleres Bruttomonatsgehalt (€)",high = "#DBD75F", low = "black") +
  labs(title = "Das Medianbruttomonatsgehalt Deutschlandweit") +
  theme_minimal() + 
  theme(legend.position = "right") + 
  theme(plot.title = element_text(size=18),
        legend.title = element_text(size=16),
        legend.text = element_text(size=14))
```












