---
title: "Non parametric tests in R"
author: "Piotr Wieczorek"
date: "2023-06-14"
output: 
  html_document: 
    theme: lumen
    toc_depth: 2
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message = FALSE)
```

## **Goal of this project**

### The aim of this project is to demonstrate various non-parametric tests using the gapminder dataset as an example. Following tests will be conducted:
<h3>1. the Wilcoxon Signed Rank test</h3>
<h3>2. the Mann U-Whitney test (otherwise known as the Wilcoxon Sum Rank test)</h3>
<h3>3. the Kruskal Wallis test (with post hoc Dunn's test)</h3>
<h3>4. Friedman's ANOVA</h3>

```{r}
library(tidyverse)
library(caret)
library(rstatix)
library(png)
library(ggpubr)
library(gapminder)
```


```{r}
df <- gapminder
```

```{r}
unique(df$year)
```

## **Wilcoxon test - 2 dependent groups**

### HO: Median life expectancy in Europe in 2002 = Median life expectancy in Europe in 2007
### Ha: Medians are not equal

```{r}
df1 <- df %>%
  filter(year %in% c(2002,2007),
         continent %in% c("Europe")) %>%
  select(1,2,3,4)
```

```{r}
head(df1)
```
```{r}
df1 %>%
  ggplot(aes(x=lifeExp))+
  geom_density()
```

### LifeExp doesn't follow normal distribution

```{r}
shapiro.test(df1$lifeExp)
```

```{r}
df1 <- df1%>%
  mutate(year = as.factor(year))
```


```{r}
df1 %>%
  group_by(year) %>%
  summarise(median_life_exp = median(lifeExp))
```
```{r}
df1 %>%
  ggplot(aes(x=year,y=lifeExp))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter()+
  labs(title = "Life Expectancy in Europe in 2002 and 2007")
```


```{r}
wilcox_test(data=df1,formula = lifeExp ~ year,paired = TRUE)
```
### Sum of all positive ranks = 0, meaning that each lifeExp in 2007 was higher than in 2002.



### Conclusion: at p-value < 0.05 there is a proof to reject the null hypothesis.



## **Mann U-Whitney test: 2 independent groups**

### hypotheses:
<h3>H0: Median life expectancy in Europe in 2007 = median life expectancy in Asia in 2007</h3>
<h3>Ha: Medians are not equal</h3>

```{r}
df2 <- gapminder
```

```{r}
df2<-df2 %>%
  filter(continent %in% c("Europe","Asia"),
         year == 2007) %>%
  mutate(year=as.factor(year)) %>%
  mutate(continent = as.factor(continent)) %>%
  select(1,2,3,4)
```



```{r}
levels(df2$year)
```
```{r}
table(df2$continent)
```
```{r}
df2$continent <- droplevels(df2$continent)
```

```{r}
table(df2$continent)
```

```{r}
df2 %>%
  ggplot(aes(x=lifeExp))+
  geom_density()+
  facet_wrap(~continent,scales="free")
```

```{r}
shapiro.test(df2[df2$continent=="Europe",]$lifeExp)
```
```{r}
shapiro.test(df2[df2$continent=="Asia",]$lifeExp)
```

```{r}
df2
```



```{r}
df2 %>%
  group_by(continent) %>%
  summarise(median_life_exp = median(lifeExp))
```

### Mann-Whitney U statistic formula



```{r,fig.width=5,fig.height=3}
ggplot() + 
    background_image(readPNG("images/mann.png"))
```


```{r}
df2 %>%
  ggplot(aes(x=continent,y=lifeExp))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter()+
  labs(title="Life Expectancy in Asia and Europe in 2007")
```


```{r}
mann_whitney <- wilcox_test(data=df2,formula = lifeExp ~ continent, paired = FALSE)
```

```{r}
mann_whitney
```




### p-value = 1.94e-05 indicates that we should reject H0. It indicates that median life expectancy for Asia and Europe is different.


## **Kruskal - Wallis - at least 3 independent groups**

### hypotheses:
<h3>H0: Median life expectancy in Europe in 2007 = Median life expectancy in Asia in 2007 = Median life expectancy in Americas in 2007</h3>
<h3>Ha: at least one pair doesn't exhibit equal medians</h3>



```{r}
df3 <- gapminder
```


```{r}
table(df3$continent,df3$year)
```


```{r}
df3<-df3 %>%
  filter(continent %in% c("Europe","Asia","Americas"),
         year == 2007) %>%
  mutate(continent = as.factor(continent),
         year = as.factor(year)) %>%
  select(2,3,4)
```


```{r}
df3$continent <- droplevels(df3$continent)
```


```{r}
table(df3$continent)
```

```{r}
df3
```


### Kruskal Wallis statistic formuala

```{r,fig.width=5,fig.height=1.5}
ggplot() + 
    background_image(readPNG("images/kruskal_wallis.png"))
```



```{r}
df3 %>%
  ggplot(aes(x=continent,y=lifeExp))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter()+
  labs(title="Life Expectancy by Continent in 2007")
```


```{r}
kruskal_test(data=df3,lifeExp~continent)
```

### Conclusion: p-value = 2.02e-05. The null hypothesis should be rejected.

## **Dunn's test - in order to evaluate differences between groups**

```{r}
dunn_test(data=df3,formula=lifeExp~continent)
```

### Significant differences in median in groups:
<h4>* Americas:Europe</h4>
<h4>* Asia:Europe</h4>

## **Friedman's ANOVA - at least 3 dependent groups**

### hypotheses:
<h3>H0: Medians across all groups are equal</h3>
<h3>Ha: at least on pair of comparisons doesn't exhibit equal median</h3>

```{r}
df4 <- gapminder
```

```{r}
df4 <- df4 %>%
  filter(continent == "Europe",
         year %in% c(1997,2002,2007)) %>%
  mutate(year = as.factor(year))
```

```{r}
table(df4$year)
```
```{r}
df4 <- df4%>%
  select(1,2,3,4)
```


```{r}
df4
```


```{r}
df4$id <- rep(seq(1:30),each=3)
```

```{r}
df4
```

### Friedman's ANOVA statistic


```{r,fig.width=6,fig.height=2}
ggplot() + 
    background_image(readPNG("images/friedman_formula.png"))
```


```{r}
df4 %>%
  group_by(year)%>%
  summarise(median_life_exp = median(lifeExp))
```

```{r}
df4 %>%
  ggplot(aes(x=year,y=lifeExp)) +
  geom_boxplot()+
  geom_jitter()+
  labs(title="Life Expectancy in Europe by Year")
```


```{r}
friedman.test(df4$lifeExp,df4$year,blocks = df4$id)
```
### Based on p-value < 0.05, we should reject the H0.



