---
title: "Exam scores"
author: "Piotr Wieczorek"
date: "2023-06-30"
output: 
  html_document: 
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
```

# **Exam scores analysis**

## Libraries

```{r}
library(readr)
library(tidyverse)
library(scales)
library(ggthemes)
library(hrbrthemes)
library(janitor)
library(caret)
library(ggpubr)
library(plyr)
```

## Theme for plotting

```{r}
th <- theme(title = element_text(size=16),
            axis.title.x = element_text(size=14),
            axis.title.y = element_text(size=14),
            axis.text.x = element_text(size=14),
            axis.text.y = element_text(size=14),
            legend.text = element_text(size=14),
            legend.title = element_text(size=14))
```

## Reading, cleaning and exploring the data

```{r}
df <- read_csv("C:/Users/piotr/Desktop/pythonfiles/exams.csv")
```

```{r}
head(df)
```
```{r}
df <- janitor::clean_names(df)
```

```{r}
head(df)
```
### Change characters to factors

```{r}
df <- df %>%
  mutate_if(is.character,as.factor)
```

### Add overall_score column

```{r}
df$overall_score <- df$math_score + df$writing_score + df$reading_score
```

### Change column name

```{r}
colnames(df)[1] <- "sex"
```

### Recode race_ethnicity factors

```{r}
df %>%
  mutate(race_ethnicity = fct_recode(race_ethnicity,"A" = "group A",
                                                    "B" = "group B",
                                                    "C" = "group C",
                                                    "D" = "group D",
                                                    "E" = "group E"))
```

### Summary for exam scores

```{r}
summary(df[,c("math_score","reading_score","writing_score","overall_score")])
```

### Plot 1 - Exam scores by exam type

#### Math score has the lowest score median, while reading score has the highest score median

```{r,fig.width=10,fig.height=6}
pivot_longer(data=df,cols=c("math_score","reading_score","writing_score"),
                            names_to = "exam_type",
                            values_to = "score") %>%
  select(c("exam_type","score")) %>%
  mutate(exam_type = fct_reorder(exam_type, score)) %>%
  ggplot(aes(x=exam_type,y=score))+
  labs(title = "Boxplots of Exam Scores by Exam Type",
       x="Exam type",
       y="Score")+
  geom_boxplot(fill = "#D5C170") + 
  annotate("text",x=0.75, y= 85, label='Median : 66.5',size = 5)+
  annotate("text",x=1.75,y=85,label='Median : 68.00',size = 5)+
  annotate("text",x=2.75, y= 85, label = "Median : 70",size = 5)+
  theme_bw()+
  th+
  ggpubr::bgcolor("#F4F2E9")
```

### Plot 2 - Exam scores by exam type distributions

```{r,fig.width=10,fig.height=6}
df %>%
  gather(c("math_score","reading_score","writing_score"),key = "key",value = "val") %>%
  select(key,val) %>%
  ggplot(aes(x=val)) + 
  geom_density(fill = "#D5C170",alpha=0.5) + 
  facet_wrap(~key,scales="free") + 
  theme_bw()+
  th+
  ggpubr::bgcolor("#F4F2E9")
```


### Median overall scores by parental level of education summary

```{r}
overall_score_stats <- ddply(df, .(parental_level_of_education), summarise, median_val = median(overall_score),
first_quantile = quantile(overall_score,0.25),
third_quantile = quantile(overall_score,0.75))
```

```{r}
overall_score_stats
```

### Figure 2 - Overall exam scores by parental level of education

```{r,fig.width=12,fig.height=8}
df %>%
  mutate(parental_level_of_education = fct_reorder(parental_level_of_education,overall_score)) %>%
  ggplot(aes(x=parental_level_of_education,y=overall_score)) + 
  labs(title="Boxplots Of Overall Score by Parental Education",
       x="Parental level of education",
       y="Overall score")+
  geom_boxplot(fill = "#D5C170")+
  geom_text(data=overall_score_stats, aes(x=parental_level_of_education, y=first_quantile + 3, label=paste("1Q =",first_quantile)),colour="black",size=5)+
  geom_text(data=overall_score_stats, aes(x=parental_level_of_education, y=median_val + 3, label=paste("2Q =",median_val)),colour="black",size=5)+
  geom_text(data=overall_score_stats, aes(x=parental_level_of_education, y=third_quantile, label=paste("3Q =",third_quantile)),colour="black",,size=5,vjust=1)+
  theme_bw()+
  th+
  ggpubr::bgcolor("#F4F2E9")
```

### Mean scores by race_ethnicity and sex

```{r}
race_ethnicity_score_stats <- ddply(df, .(race_ethnicity,sex), summarise, median_val = median(overall_score),
first_quantile = quantile(overall_score,0.25),
third_quantile = quantile(overall_score,0.75))
```


```{r}
race_ethnicity_score_stats
```

### Figure 2 - Overall exam scores by parental level of education

#### Overall, female students perform better than male students, especially in group A. Interestingly, group E exhibits the best overall score results, although their parents have the least bachelor's and master's degrees.

```{r,fig.width=12,fig.height=8}
df %>%
  mutate(race_ethnicity = fct_reorder(race_ethnicity,overall_score)) %>%
  ggplot(aes(x=race_ethnicity,y=overall_score,fill=sex)) +
  geom_boxplot()+
  scale_fill_manual(values=c("orange","#E1E285"))+
  geom_text(data = race_ethnicity_score_stats, 
            aes(x=race_ethnicity,
                y=80,
                label=paste("1Q =",first_quantile)),position=position_dodge(.9), size = 4.5)+
  geom_text(data = race_ethnicity_score_stats,
            aes(x=race_ethnicity,
                y=65,
                label=paste("2Q =",median_val)),
            position=position_dodge(.9), size = 4.5) +
  geom_text(data = race_ethnicity_score_stats,
            aes(x=race_ethnicity,
                y=50,
                label=paste("3Q =",third_quantile)),
            position=position_dodge(.9), size = 4.5)+
  labs(title = "Overall Score by Ethnicity Group and Sex",
       x = "Race/ethnicity",
       y = "Overall score")+
  theme_bw()+
  th+
  ggpubr::bgcolor("#F4F2E9")
```

### Number of parental level of education categories by ethnicity group

```{r}
janitor::tabyl(df,parental_level_of_education,race_ethnicity)
```

### Conclusions:

<h4>1. Female students perform better than male students</h4>
<h4>2. Group E shows the highest median of the overall score when Group B shows the lowest median of the overall score</h4>
<h4>3. Median of math exam results is the lowest while median of reading exam results is the highest</h4>
<h4>4. Median of overall score results was the highest for students whose parents have master's degrees and the lowest for students whose parents have "some high school" education</h4>





