---
title: "Var Model - Predicting Number of Dwellings for which planning permission has been granted or an application has been made with a construction project"
author: "Piotr Wieczorek"
date: "2023-12-07"
output: 
  html_document: 
    toc: yes
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

```{r}
rm(list=ls())
```

## Libraries

```{r}
library(readxl)
library(tidyverse)
library(lubridate)
library(zoo)
library(patchwork)
library(tseries)
library(forecast)
library(vars)
library(Metrics)
library(lmtest)
```

## Reading and cleaning the data

```{r}
df <- read_excel('Mieszkania.xlsx')
```



### Changing colnames



```{r}
colnames(df) <- c("time",
"dwellings_planning_permission_granted",
"general_business_climate_index_construction",
"reference_interest_rate",
"inflation_rate"
)
```


```{r}
head(df)
```

### Mutating columns

```{r}
df <- df %>%
  mutate(across(c(2:5),~str_replace_all(.,",","."))) %>%
  mutate(across(c(2:5),~as.numeric(.))) %>%
  mutate(inflation_rate = inflation_rate - 100)
```


```{r}
df <- df %>%
  mutate(time = as.factor(time))
```

```{r}
df$time <- zoo::as.yearqtr(df$time,format="%Y-%qQ")
```


```{r}
head(df)
```

### Creating time series objects

```{r}
ts_dwellings_planning_permission_granted <- ts(df$dwellings_planning_permission_granted,start=c(2000,1),frequency = 4)
ts_general_business_climate_index_construction <- ts(df$general_business_climate_index_construction,start=c(2000,1),frequency = 4)
#ts_reference_interest_rate <- ts(df$reference_interest_rate, start=c(2000,1),frequency = 4)
ts_inflation_rate <- ts(df$inflation_rate,start=c(2000,1),frequency = 4)
```


```{r}
th <- theme(plot.title=element_text(size=18),
            axis.title.x = element_text(size=16),
            axis.title.y = element_text(size=16),
            axis.text.x = element_text(size=14),
            axis.text.y = element_text(size=14),
            legend.title = element_text(size=18),
            legend.text = element_text(size=14))
```



## Exploratory analysis

### Plotting the data

```{r,fig.width=12,fig.height=6}
p1 <- df %>%
  ggplot(aes(x=time,y=dwellings_planning_permission_granted)) + 
  geom_line(lwd=1.4)+ labs(y="",title="Permission Granted for New Dwellings") + theme_minimal()  + th



p2 <- df %>%
  ggplot(aes(x=time,y=general_business_climate_index_construction))  +
  geom_line(lwd=1.4) +  labs(y="",title="General Construction Business Climate Index") + theme_minimal()+ th

p3<- df %>%
  ggplot(aes(x=time , y = inflation_rate)) + 
  geom_line(lwd=1.4) + labs(y="",title = "Inflation Rate") + theme_minimal() +  th

(p1+p2)/(p3)
```



### ADF tests


```{r}
adf.test(ts_dwellings_planning_permission_granted)
```


```{r}
adf.test(ts_general_business_climate_index_construction)
```

```{r}
adf.test(ts_inflation_rate)
```


## Modeling

### Binding times series data by columns

```{r}
df2 <- cbind(ts_dwellings_planning_permission_granted,ts_general_business_climate_index_construction,ts_inflation_rate)
```


### Selecting optimal lag for the model


```{r}
VARselect(df2,lag.max = 4)
```



### Model formula


```{r}
estim <- VAR(df2,p=4,type="const")
```


### Summary

```{r}
summary(estim$varresult$ts_dwellings_planning_permission_granted)
```

### Calculating MAE and MAPE

```{r}
mod_res <- estim$varresult$ts_dwellings_planning_permission_granted$residuals
mod_fit <- estim$varresult$ts_dwellings_planning_permission_granted$fitted.values
```

```{r}
mean(abs(mod_res))
```

```{r}
actual_values <- ts_dwellings_planning_permission_granted[5:95]
predicted_values <- estim$varresult$ts_dwellings_planning_permission_granted$fitted.values
```

```{r}
mae(actual_values,predicted_values)
```
```{r}
mape(actual_values,predicted_values) #mozna policzyc mape dlatego ze w zadnym z actual_values nie ma 0
```

### Casuality tests


####  Pr(>F) < 0.05 implies that there is statistical evidence to support the idea that past values of ts_inflation_rate Granger cause ts_dwellings_planning_permission_granted.

```{r}
lmtest::grangertest(ts_dwellings_planning_permission_granted~ts_inflation_rate,order=4)
```

####  Pr(>F) < 0.05 implies that there is statistical evidence to support the idea that past values of ts_general_business_climate_index_construction Granger cause ts_dwellings_planning_permission_granted.


```{r}
lmtest::grangertest(ts_dwellings_planning_permission_granted~ts_general_business_climate_index_construction,order=4)
```


### Forecast error variance decomposition

#### Both inflation rate and general business climate index for construction have an impact on explaining forecast error of the ts_dwellings_planning_permission_granted variable


```{r,fig.width=10,fig.height=10}
plot(fevd(estim, n.ahead =8))
```


### Impulse response function


#### There is an increase of ts_dwellings_planning_permission_granted as a reaction to the sudden (positive) shock in ts_general_business_climate_index_construction


```{r,fig.width=12,fig.height=8}
plot(irf(estim,impulse="ts_general_business_climate_index_construction",response = "ts_dwellings_planning_permission_granted",n.ahead = 4),main="Reaction of ts_dwellings_planning_permission_granted on ts_general_business_climate_index_construction impulse",cex=1.25)
```



#### There is a decrease of ts_dwellings_planning_permission_granted as a reaction to the sudden (positive) shock in ts_inflation_rate


```{r,fig.width=12,fig.height=8}
plot(irf(estim,impulse="ts_inflation_rate",response = "ts_dwellings_planning_permission_granted",n.ahead = 4),main="Reaction of ts_dwellings_planning_permission_granted on ts_inflation_rate impulse",cex=1.25)
```



```{r,fig.width=10,fig.height=10}
plot(predict(estim,n.ahead = 12))
```








