---
title: "Crime Exploratory Analysis"
author: "Piotr Wieczorek"
date: "2023-11-27"
output: 
  html_document: 
    toc: yes
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

## Libraries

```{r}
library(tidyverse)
library(caret)
library(naniar)
library(janitor)
library(lubridate)
library(scales)
library(viridis)
```

```{r}
rm(list=ls())
```

## Reading and cleaning the data

```{r}
df <- read.csv("https://query.data.world/s/fie774raw23pzspuhjqwpbpmjtogem?dws=00000", header=TRUE, stringsAsFactors=FALSE);
```

### Cleaning colnames

```{r}
df <- janitor::clean_names(df)
```

### Unselecting and mutating columns
```{r}
df <- df %>%
  select(-c("location_1","post","crime_code")) %>%
  mutate_at(c("crime_date"), lubridate::mdy) %>%
  mutate(across(where(is.character),~str_to_upper(.))) %>%
  mutate(across(where(is.character),as.factor)) %>%
  mutate(inside_outside = fct_recode(inside_outside,
                                     OUTSIDE = "O",
                                     INSIDE = "I"))
```

### Creating new column (crime_hour)

```{r}
df$crime_hour <- str_sub(df$crime_time,start=1,end=2)
df$crime_hour <- as.numeric(df$crime_hour)
```


```{r}
table(df$crime_hour)
```

```{r}
df$crime_hour <- ifelse(df$crime_hour == 24, 0,df$crime_hour)
```

```{r}
table(df$crime_hour)
```


```{r}
head(df)
```

### Creating new column (day_of_the_week)

```{r}
df$day_of_week <- lubridate::wday(df$crime_date,label=TRUE,abbr=FALSE)
```



```{r}
df <- df %>%
  mutate(across(c("day_of_week","crime_time"), as.character))
```

### Exploring NA values

```{r}
colSums(is.na(df))
```

### Exploring blank values

```{r}
colSums(df=="")
```

### Replacing blank values with NA

```{r}
df <- replace(df, df=='', NA)
```


```{r}
colSums(is.na(df))
```

### NA values report

```{r}
naniar::miss_var_summary(df)
```

### Unselecting total_indidents column

#### This column contains only one value which is "1" so there is no reason to keep it.

```{r}
table(df$total_incidents)
```
```{r}
df <- df %>%
  select(-c("total_incidents"))
```

```{r}
head(df)
```

### Filtering out NA values from selected columns

```{r}
df <- df[complete.cases(df[,c("inside_outside","neighborhood","location","district")]),]
```

### Replacing NA values in "weapon" column with a specified category

#### It seems that no entity provided in the "weapon" column means that no weapon was used but it was arbitrary to make this conclusion 

```{r}
df <- df %>%
  mutate(weapon = as.character(weapon)) %>%
  mutate(across(c("weapon"),~replace_na(.,"NO REGISTERED WEAPON")))
```

```{r}
head(df)
```


```{r}
miss_var_summary(df)
```

```{r}
df <- df %>%
  select(c(1,2,9,everything()))
```

```{r}
head(df)
```

## Data visualization

### Crime frequency by hour

```{r,fig.width=10}
df %>%
  group_by(crime_hour) %>%
  summarise(freq = n()) %>%
  ggplot(aes(x=crime_hour,y=freq)) + 
  geom_bar(stat="identity",color="black",fill="#6080B0") + 
  geom_text(aes(label=freq),vjust=-0.25,size=3.25) + 
  scale_x_continuous(breaks=seq(0,23,by=1)) + 
  labs(title="Crime Frequency by Hour")+
  theme_light()
```

### Changing the level order for the day_of_week column


```{r}
df$day_of_week <- as.factor(df$day_of_week)
levels(df$day_of_week)
```


```{r}
df$day_of_week <- factor(df$day_of_week, levels=c('Monday', 'Tuesday', 'Wednesday',
                                                  'Thursday', 'Friday','Saturday','Sunday'))
```

### Crime frequency by hour and day


```{r,fig.width=10}
df %>%
  group_by(crime_hour,day_of_week) %>%
  summarise(freq = n()) %>%
  ggplot(aes(x=crime_hour,y=freq,fill=day_of_week)) + 
  geom_bar(stat="identity",
           position = "stack",
           color="black") + 
  #geom_text(aes(label=freq),vjust=-0.25,size=3.25) + 
  scale_x_continuous(breaks=seq(0,23,by=1)) +
  scale_fill_viridis(discrete = TRUE)+
  labs(title="Crime Frequency by Hour and Day",subtitle = "2011/01/01 - 2016/11/12")+
  theme_light()
```



### Crime Frequency by description


```{r,fig.width=8}
df %>%
  group_by(description) %>%
  summarise(freq = n()) %>%
  arrange(freq) %>%
  mutate(description=factor(description, levels=description)) %>%
  ggplot(aes(x=description, y= freq)) + 
  geom_segment( aes(xend=description, yend=0)) +
    geom_point( size=2, color="orange") +
    coord_flip() +
    labs(title = "Crime Frequency by Description",
         subtitle = "2011/01/01 - 2016/11/12",
         y="") +
    scale_y_continuous(label=comma,breaks=seq(0,60000,by=10000)) + 
    theme_bw() +
    xlab("")
```



### Crime frequency by hour and day

```{r,fig.width=10}
for (day_name in levels(df$day_of_week)) {
  p <- df %>%
  group_by(crime_hour,day_of_week) %>%
  summarise(freq = n()) %>%
  filter(day_of_week == day_name) %>%
  ggplot(aes(x=crime_hour,y=freq)) + 
  geom_bar(stat="identity",color="black",fill="#6080B0") + 
  geom_text(aes(label=freq),vjust=-0.25,size=3.25) + 
  scale_x_continuous(breaks=seq(0,23,by=1)) + 
  #facet_wrap(~day_of_week,scales="free") + 
  labs(title=paste("Crime Frequency by Hour in",day_name))+
  theme_light()
print(p)
}
```

### Fixing the "district" column

```{r}
df$district <- droplevels(df$district)
table(df$district)
```

#### Collapsing the factors


```{r}
df <- df[df$district != "GAY STREET",]
df <- df %>%
  mutate(district = fct_collapse(district,
                                 "NORTH" = c("NORTHEASTERN","NORTHERN","NORTHWESTERN"),
                                 "SOUTH" = c("SOUTHEASTERN","SOUTHERN","SOUTHWESTERN")))
```

```{r}
df$district <- droplevels(df$district)
table(df$district)
```

### Crime frequency by district

```{r}
df %>%
  group_by(district) %>%
  summarise(freq = n())  %>%
  arrange(freq) %>%
  mutate(district = factor(district, levels= district)) %>%
  ggplot(aes(x=district,y=freq)) + 
  geom_segment(aes(xend=district,yend=0)) + 
  geom_point(size=2,color="orange") + 
  coord_flip() + 
  labs(title = "Crime Frequency by District",
         subtitle = "2011/01/01 - 2016/11/12",
         y="") + 
    theme_bw() +
    xlab("")
```

### Weapon shape (for commited crimes) by district

```{r}
df %>%
  group_by(district,weapon) %>%
  summarise(freq = n()) %>%
  mutate(district = fct_reorder(district,freq)) %>%
  ggplot(aes(x=district,y=freq,fill=weapon)) + 
  geom_bar(stat="identity",position="fill",color="black") + 
  scale_fill_viridis(discrete = TRUE) + 
  labs(y="%",title = "Weapon share by District")+
  theme_light()
```

