---
title: "Predicting median income in USA with the usage of spatially lagged linear model"
output: 
  html_document: 
    toc: true
    theme: united
date: "2024-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=F,message = F)
```


```{r}
rm(list=ls())
```



## Libraries


```{r}
library(tidyverse)
library(janitor)
library(skimr)
library(sf)
library(sfdep)
library(spdep)
library(spatial)
library(spatialreg)
library(ggmap)
library(mapproj)
library(maps)
library(readxl)
library(scales)
library(gridExtra)
library(cowplot)
library(png)
library(ggpubr)
```

## Reading the data

```{r}
geo_data <- st_read("us_states_coordinates.json")
```


```{r}
invalid_geometries <- geo_data %>% filter(!st_is_valid(.))
invalid_geometries
```
```{r}
# Attempt to make invalid geometries valid
geo_data <- geo_data %>% mutate(geometry = st_make_valid(geometry))
```



```{r}
geo_data$name
```



```{r}
income_data <- read_excel("us_states_income.xlsx")
```




```{r}
urban_data <- read_excel("us_urbanization.xlsx")
density_data <- read_excel("us_density.xlsx")
age_data <- read_excel("us_median_age.xlsx")
degree_data <- read_excel("us_states_bachelor.xlsx")
```


```{r}
urban_data <- urban_data %>%
  mutate(urbanization = as.numeric(urbanization))

density_data <- density_data %>%
  mutate(population_density = as.numeric(population_density))

age_data <- age_data %>%
  mutate(median_age = as.numeric(median_age))

degree_data <- degree_data %>%
  mutate(population_over_25 = as.numeric(population_over_25))
```


```{r}
income_data <- income_data %>%
  inner_join(urban_data,by="us_state") %>% 
  inner_join(density_data,by="us_state") %>% 
  inner_join(age_data,by="us_state") %>% 
  inner_join(degree_data,by="us_state")
```



```{r}
income_data <- income_data %>%
  rename("name"="us_state")

income_data <- geo_data %>%
  inner_join(income_data,by="name")
```



```{r}
class(income_data)
```


## Creating neighbour list object


```{r}
nb <- st_contiguity(income_data)
```


```{r}
nb
```



```{r}
nb[1:3] #first observation has neighbors with observation 8,9,22,38 and so on..
```

## Compute spatial weights (row-standarized)

#### Row-standarization means that each element in the weight matrix gets divided by the sum within each row

```{r}
wt <- st_weights(nb) #spatial weights (row standardized)
```


```{r}
wt
```

## Spatial lag
#### Spatial lag is computed as sum of multiplication of each neighbor row-standarized weight and its value for the variable of interest.

```{r}
x <- income_data$median_income
```


```{r}
st_lag(x,nb,wt)
```

#### For example, let’s take a look on the spatial lag for the first observation in the data set:
#### The first observation is Alabama, which has neighbors with indexes 8, 9, 22, 38 which are:
* Florida 
* Georgia
* Mississippi
* Tennessee

#### Since Alabama has 4 neighbors, all of them will have the weight of 0.25 in the weight matrix
#### These are the median incomes values for the neighbors:

* Florida 65370 
* Georgia 67630 
* Mississippi 48610 
* Tennessee 65380


#### To calculate the spatial lag for Alabama, we will have to multiply the weights (all are equal to 0.25) with the median incomes

```{r}
ij <- nb[[1]]
wij <- wt[[1]]
xij <- x[ij]
```

```{r}
ij #Neighbors indexes
```

```{r}
wij #Weights 
```

```{r}
xij #Median incomes in Florida, Georgia, Mississippi, Tennessee respectively
```

```{r}
as.tibble(income_data) %>%
  filter(name %in% c("Florida","Georgia","Mississippi","Tennessee")) %>%
  select(c(name,median_income))
```


#### First spatial lag value

```{r}
sum(xij*wij) #Spatial lag value for Alabama
```


## Creating a data frame with neighbors, weights and spatial lag values


```{r}
income_lags <- income_data %>%
  transmute(nb=st_contiguity(geometry),
            wt = st_weights(nb),
            income_lag = st_lag(median_income,nb,wt))
```

```{r}
income_lags %>%
  as_tibble() %>%
  head()
```



## Comparing map plots - Median income in the US vs Spatially lagged median income in the US

#### Note that the second plot (spatially lagged median income) is much more clustered as it should be

```{r,fig.width=10,fig.height=8}
 

plot1<- ggplot(income_data,aes(fill=median_income,
                       label=median_income)) +
  geom_sf()+
 #geom_sf_label(fun.geometry = sf::st_centroid,size=3,fill="white") + 
  scale_fill_gradient(name = "Median Income by US State", low = "#C4E1C4", high = "#008802") +
  labs(title="Median Income in USA",
       x="",
       y="") +
  theme(plot.title = element_text(size=20,color="black"),
        legend.title = element_text(size=16,color="black"),
        legend.text = element_text(size=16,color="black"),
        legend.position = "bottom",
        plot.caption=element_text(size=16),
        text = element_text(family = "Times New Roman")) + theme_minimal()
 

legend <- get_legend(plot1)

plot2 <-income_lags %>%
  ggplot(aes(fill=income_lag)) + 
  geom_sf(color="black",lwd=0.1) + 
 #geom_sf_label(fun.geometry = sf::st_centroid,size=3,fill="white") + 
  scale_fill_gradient(name = "Median Income by US State", low = "#C4E1C4", high = "#008802",limits= range(income_data$median_income)) +
  labs(title="Median Income in USA - Spatial lags",
       x="",
       y="") +
  theme(plot.title = element_text(size=20,color="black"),
        legend.title = element_text(size=16,color="black"),
        legend.text = element_text(size=16,color="black"),
        legend.position = "bottom",
        plot.caption=element_text(size=16),
        text = element_text(family = "Times New Roman")) + theme_minimal()

  
plot1 <- plot1 +theme(legend.position = "none")
plot2 <- plot2 +theme(legend.position = "none")
 

combined_plot <- plot_grid(plot1,plot2,ncol=1)

final_plot <- plot_grid(
  combined_plot, 
  legend, 
  ncol = 2,
  rel_widths = c(2,1))


print(final_plot)

```


## Creating a linear model WITHOUT special dependence


```{r}
income_data %>%
  as_tibble() %>%
  head()
```


```{r}
mod1 <- lm(data = income_data, formula = median_income ~ urbanization + population_density + median_age  + population_over_25)
```

```{r}
summary(mod1)
```


#### Population density is highly insignifant so let's remove this variable from the model


```{r}
mod1 <- lm(data = income_data, formula = median_income ~ urbanization + median_age  + population_over_25)
```

```{r}
summary(mod1)
```

## Creating a linear model WITH spatial dependence

#### Moran test for regression residuals
#### H0: No spatial dependence in regression residuals
#### Since p-value < 0.05, it should be concluded that there exists a spatial dependence within linear regression’s residuals, which means that a spatial linear model would be better than “default” linear regression model


```{r}
lm.morantest(mod1,listw = nb2listw(nb))
```

## Creating SAR spatial lag model 

#### Formula


```{r,fig.width=6,fig.height=2}
ggplot() + 
    background_image(readPNG("sar_formula.png")) + 
  labs(caption = "Source:https://web.sgh.waw.pl/~atoroj/ekonometria_przestrzenna/5_single_source_GIS_polygons_EN.pdf")
```

##### The formula essentially means: Take the sum of multiplication of weights and Y values for each neighbor, scale with with Rho and add the X variables multiplied by their coefficients



```{r}
mod_spatial <- lagsarlm(formula = median_income ~ urbanization + median_age  + population_over_25,
    data = income_data,
    listw = nb2listw(nb))
```


```{r}
summary(mod_spatial,Nagelkerke = TRUE)
```
#### Nagelkerke pseudo-R-squared is slighly higher than R^2 (0.81225 compared to 0.7757 in the OLS model)

#### All variables in the SAR model are significant.

## Direct and indirect effect


#### Due to the feedback loop (global spatial lag was used), it is needed to compute the direct (the impact of a particular variable on the Y variable in a particular neighbor) and the indirect effect (the impact of a particular variable in the neighbors of a particular neighbor on the Y value of that neighbor)


```{r}
impacts(mod_spatial,listw = nb2listw(nb))
```

### Interpretation regarding the direct effect

#### If urbanization increases by 1 unit, median income will increase by 135, holding other variables constant, in the same spatial unit.
#### If median_age increases by 1 unit, median incomee will decrease by -1298.6780, holding other variables constant in the same spatial unit
#### If population_over_25 (% of people being at least 25 years old with a bachelor's degree or higher) increases by 1 unit, median income will increase by 1542.0131


### Interpretation regarding the indirect effect


#### If urbanization increases by 1 unit, median income will increase by 42.37259 in neighboring units
#### If median_age increases by 1 unit, median income will decrease by -406.92602 in neighboring units
#### If population_over_25 (% of people being at least 25 years old with a bachelor's degree or higher) increases by 1 unit, median income will increase by 483.16896 in neighboring units


