---
title: "Using Prophet to predict newly registered vehicles in Poland"
author: "Piotr Wieczorek"
date: "2023-12-16"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

```{r}
rm(list=ls())
```

## Libraries

```{r}
library(tidyverse)
library(tidymodels)
library(prophet)
library(readxl)
library(scales)
library(gridExtra)
```


## Reading and cleaning the data


```{r}
df <- read_excel("time_series_vehicles_data.xlsx")
```

```{r}
colnames(df) <- c("ds","y")
```

```{r}
df[c(99,100),2] <- df[98,2]
```

```{r}
head(df)
```

```{r}
tail(df)
```

```{r,fig.width=10,fig.height=5}
df %>%
  ggplot(aes(x=ds,y=y)) + 
  geom_line() + 
  scale_y_continuous(labels=scales::comma) + 
  labs(x="Date",y="Newly registered vehicles",title = "Newly Registered Vehicles in Poland") + 
  theme_bw(base_size = 15)
```

## Modeling time series with prophet

### Seasonality plot

#### Conclusions regarding seasonality

##### There is yearly seasonality. 

##### There are less newly registered vehicles compared to the overall trend in following time stamps:

* From Januar to mid Februar
* From the beginning of May to to the beginning of June there is large drop. There are less newly registered vehicles compared to the overall trend
* From the beginning of August to the beginning of September
* From the beginning of October to the beginning of December
*


##### There are more newly registered vehicles compared to the overall trend in following time stamps:
* From mid February to the beginning of May
* From the beginning of June to the beginning of August (especially in June)
* From the beginning of September to the beginning of October
* From the beginning of December to late December

```{r,fig.show='hide'}

model <- prophet(df=df %>% arrange(ds),seasonality.mode = "additive")

future <- make_future_dataframe(model, periods = 24, freq = "month")

forecast <- predict(model,future)

plot_list <- prophet_plot_components(model,forecast)
```


```{r,fig.width=10,fig.height=5}
  
plot_list[[2]] +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") + 
  scale_x_datetime(date_breaks = "month",date_labels = "%b") + 
  scale_y_continuous(labels=comma)+
  labs(title = "Seasonal Component",x="Time",y="Trend") + 
  theme_bw(base_size = 14) 

```

### Trend plot

#### Conclusions regarding trend

##### Prophet uses piecewise trend. It's clear that the variable (newly registered vehicles in Poland) has been growing quite linearly from 2012 to ~2019, and has been declining since then also in an approximately linear fashion

```{r,fig.width=10,fig.height=5}
  
  plot_list[[1]] + 
  scale_x_datetime(date_breaks = "year",date_labels = "%Y") + 
  scale_y_continuous(labels=comma) + 
  labs(title = "Trend Component",x="Time",y="Trend") + 
  theme_bw(base_size = 14)

```


```{r}
paste("Seasonality mode:",model$seasonality.mode)
paste("Growth mode:",model$growth)
```


#### Points where the trend changes


```{r,fig.width=10,fig.height=5}
plot(model,forecast) + 
  add_changepoints_to_plot(model) + 
  scale_x_datetime(date_breaks = "year",date_labels = "%Y") + 
  scale_y_continuous(labels=scales::comma) + 
  theme_bw(base_size = 14)
```

### Interactive prediction plot

```{r,fig.width=10,fig.height=5}
dyplot.prophet(model,forecast,main="Prophet Prediction") 
```


### MAE and MAPE metrics


```{r}
# Calculate MAE
mae <- mean(abs(forecast$yhat - df$y))
print(paste("Mean Absolute Error (MAE):", mae))
```
 

```{r}
# Calculate MAPE
mape <- mean(abs((df$y - forecast$yhat) / df$y)) * 100
print(paste0("Mean Absolute Percentage Error (MAPE): ", round(mape,2),"%"))
```

















