---
title: "Time Series Analysis - Newly Registered Vehicles in Poland"
author: "Piotr Wieczorek"
date: "2023-06-23"
output: 
  html_document: 
    toc: yes
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

## **1 Goal of this project**

#### The goal of the analysis is to make a forecast of the time series presenting data on number of vehicles registered in Poland for the first time (from 01.01.2012 to 01.04.2023) for the period **01.05.2023 - 01.04.2025**

## **2 Libraries**

```{r}
library(dplyr)
library(tidyverse)
library(tseries)
library(GGally)
library(forecast)
library(readxl)
library(lubridate)
library(scales)
library(patchwork)
library(kableExtra)
library(ggpubr)
library(png)
library(jpeg)
```

## **3 Reading and preparing the data**

```{r}
df <- read_excel("time_series_vehicles_data.xlsx")
```

```{r}
df$t <- lubridate::ymd(df$t)
class(df$t)
```

```{r}
colnames(df)[2] <- "Newly_registered_vehicles"
```

```{r}
head(df)
```

### **3.1 Data correction - the beginning of the COVID-19 pandemic**

#### As the beginning of the COVID-19 pandemic (March, April 2020) led to an unpredictable standstill, the values of newly registered vehicles in Poland for these two periods have been converted to the February 2020 value.

```{r}
df[c(98:100),1:2]
```

```{r}
df[c(99,100),2] <- df[98,2]
```

```{r}
df[98:100,1:2]
```

```{r}
df_ts <- ts(df$Newly_registered_vehicles,start = 2012,frequency = 12)
```

```{r}
df_ts
```

#### Setting up the theme 

```{r}
th <- theme(
        axis.text.x = element_text(size=12),
        axis.text.y = element_text(size=12),
        title = element_text(size=16),
        axis.title.x = element_text(size=12),
        axis.title.y = element_text(size=12))
```

### **3.2 Base plot - newly registered vehicles in Poland from 01.01.2012 to 01.04.2023**

```{r,fig.width=10,fig.height=6,fig.align='center'}
autoplot(df_ts,main = "Newly Registered Vehicles in Poland by Year",xlab = "Year",ylab="Newly registered vehicles") + th
```

## **4 Other datasets**

#### Other datasets were also included in order to examine whether or not there is a correlation between these variables and the number of newly registered vehicles in Poland

### **4.1 Inflation in the tranport sector data (regarding Poland)**
```{r}
df2 <- read_excel("transport_inflation.xlsx")
```

```{r}
head(df2)
```

```{r}
df2 <- df2 %>%
mutate(transport_inflation = as.numeric(transport_inflation))
```

```{r}
df2$t <- lubridate::ymd(df2$t)
class(df2$t)
```

```{r}
df2_ts <- ts(df2$transport_inflation,start = 2012,frequency = 12)
```

```{r}
df2_ts
```

### **4.2 Salary data (regarding Poland)**

```{r}
df3 <- read_excel("average_real_salary.xlsx")
```


```{r}
head(df3)
```


```{r}
df3 <- df3 %>%
mutate(Average_monthly_real_gross_salary_business_sector = as.numeric(Average_monthly_real_gross_salary_business_sector))
```

```{r}
df3$t <- lubridate::ymd(df3$t)
class(df3$t)
```

```{r}
head(df3)
```

```{r}
df3_ts <- ts(df3$Average_monthly_real_gross_salary_business_sector,start = 2012,frequency = 12)
```

```{r}
df3_ts
```

### **4.3 Unemployment data (regarding Poland)**

```{r}
df4 <- read_excel("unemployment.xlsx")
```

```{r}
df4$t <- lubridate::ymd(df4$t)
class(df4$t)
```


```{r}
head(df4)
```

```{r}
df4_ts <- ts(df4$unemployment_rate,start = 2012,frequency = 12)
```

```{r}
df4_ts
```

## **5 Examining the correlation between newly registered vehicles and other variables**


```{r,fig.width=16,fig.height=8,fig.align='center'}
autoplot(df_ts,main="Newly Registered Vehicles in Poland") + th + autoplot(df2_ts,main = "Transport Inflation in Poland") + th +
autoplot(df3_ts,main="Average Monthly Real Salary Growth (Private Companies)") + th +
autoplot(df4_ts,main="Uneployment Rate in Poland") + th
```

#### Merging the data to include all variables in one dataframe

```{r}
df_merged <- merge(df,df2,by="t")
```


```{r}
df_merged <- merge(df_merged,df3,by="t")
```

```{r}
df_merged <- merge(df_merged,df4,by="t")
```

```{r}
head(df_merged)
```

#### The additional variables are not normally distributed therefore the correlation was measured with the spearman rank correlation

```{r,fig.width=12,fig.height=6,fig.align='center'}
df_merged %>%
gather(c(2:5),key="key",value="val") %>%
ggplot(aes(x=val))+
geom_density()+
facet_wrap(~key,scales="free") + 
theme_bw() + 
th
```

#### Shapiro wilk test results

```{r}
shapiro.test(df_merged$Average_monthly_real_gross_salary_business_sector)
```

```{r}
shapiro.test(df_merged$transport_inflation)
```

```{r}
shapiro.test(df_merged$unemployment_rate)
```

### **5.1 Correlation between newly registered vehicles in Poland and inflation in the transport sector in Poland**

```{r}
cor(df_merged$Newly_registered_vehicles,df_merged$transport_inflation,method="spearman")
```

### **5.2 Correlation between newly registered vehicles in Poland and average monthly real gross salary in private sector in Poland**

```{r}
cor(df_merged$Newly_registered_vehicles,df_merged$Average_monthly_real_gross_salary_business_sector,method="spearman")
```

### **5.3 Correlation between newly registered vehicles in Poland and unemployment rate in Poland**

```{r}
cor(df_merged$Newly_registered_vehicles,df_merged$unemployment_rate,method="spearman")
```

## **5.4 Conclusions regarding correlation**

<h4>1. Inflation in the transport sector exhibits almost no influence of the number of newly registered vehicles in Poland. It indicates that Poles don't pay too much attention to the inflation rate (regarding transport) when it comes to buying new vehicles. Perhaps consumers presume that the vehicles (especially cars) are going to be even more expensive in the future</h4>

<h4>2. There is a positive correlation between newly registered vehicles in Poland and Average monthly real gross salary in private sector in Poland</h4>

<h4>3. There is a negative correlation between newly registered vehicles in Poland and the unemployment_rate in Poland</h4>

## **6 Proper time series analysis part - newly registered vehicles in Poland**

### **6.1 Stationarity exploration**

#### In order for the time series data to be stationary, it should have:
<h4> 1. Equal mean (which indicates no trend)</h4>
<h4> 2. Equal variance (which indicates no seasonality)</h4>
<h4> 3. If the time series data was divided into multiple and equal parts, technically it should exhibit very similar distributions</h4>
<h4> 4. Low p-value (lower than alpha) from the ADF (Augmented-Dicker-Fuller) test, where H0 states non-stationarity</h4>

#### Based on the ACF plot:

<h4>1. The plot exhibits trend and seasonality, because the "spikes" are high at the beginning, and then quickly decrease in a cyclical movement</h4>
<h4>2. All of the "spikes" are out of the insignificance area which indicates high auto-correlation within the data</h4>

```{r}
ggAcf(df_ts) + ggtitle("ACF Plot") + theme_bw() + th
```


#### The PACF plot pictures significant correlation between Y at t=0 and Y at t=1,2,3,8,9,11,12,13,14

```{r}
ggPacf(df_ts) + ggtitle("PACF Plot") + theme_bw() + th
```


#### ADF Test:

<h4>H0: Non-stationary series</h4>
<h4>H1: Stationary series</h4>

```{r}
adf.test(df_ts,k=12)
```


#### The data is non-stationary because:

<h4>1. It exhibits trend (it's better pictured with the usage of the time series decomposition below)</h4>
<h4>2. It exhibits seasonality (it's better pictured with the usage of the time series decomposition below)</h4>
<h4>3. It exhibits strong auto-correlation between Yt and its lags</h4>
<h4>4. The p-value from the ADF test transcends the alpha value (0.05) and therefore there is no proof for rejecting the H0</h4>

## **7 Time series decomposition**

#### Based on the decomposition:

<h4>1. Data exhibits trend</h4>
<h4>2. Data exhibits seasonality (every 12 months)</h4>
<h4>3. What is left after taking the trend and seasonal component appears to be random (mean = 0 and no clear patterns)</h4>

```{r}
fit <- decompose(df_ts)
```

```{r}
p1 <- autoplot(fit$trend, main = "Trend Component") + theme_bw()
p2 <- autoplot(fit$seasonal, main = "Seasonal Component") + theme_bw() 
p3 <- autoplot(fit$random, main ="Random Component" ) + theme_bw()
```

```{r,fig.width=10,fig.height=8,fig.align='center'}
(p1 + th)/(p2 + th)/(p3 + th)
```


## **8 Seasonality exploration**

#### Based on the plot below:


<h4>1. There was more newly registered vehicles (on average) in months:</h4>
<h4>* March</h4>
<h4>* June</h4>
<h4>* July</h4>


<h4>2. There was fewer newly registered vehicles (on average) in months:</h4>
<h4>* January</h4>
<h4>* February</h4>
<h4>* November</h4>
<h4>* December</h4>

```{r,fig.width=10,fig.height=6,fig.align='center'}
ggsubseriesplot(df_ts,main = "Mean of Newly Registered Vehicles By Month") + 
  scale_y_continuous(breaks = seq(70000,190000,20000), labels = comma_format()) + labs(y="Newly registered vehicles",x="Month") + theme_bw() + 
  th
```


## **9 The base data, its trend and seasonality in one graph**

```{r,fig.width=10,fig.height=6,fig.align='center'}
autoplot(df_ts,series="Base data") +
  autolayer(trendcycle(fit), series="Trend") +
  autolayer(seasadj(fit),series="Seasonally adjusted data")+
  scale_y_continuous(breaks = seq(round(min(df_ts),-4),round(max(df_ts),-4),by=20000),
                     labels = comma_format())+
  scale_x_continuous(breaks=seq(min(time(df_ts)),max(time(df_ts))+1,by=1)) +
  scale_colour_manual(values=c("grey","blue","red"),
             breaks=c("Base data","Trend","Seasonally adjusted data"))+
  labs(y="Newly registered vehicles",
       x="Year",
       title="Base Data, Trend and Seasonally adjusted data")+
  theme_bw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(size=14),
        axis.text.y = element_text(size=14),
        title = element_text(size=18),
        legend.title = element_text(size=18),
        legend.text = element_text(size=14),
        axis.title.x = element_text(size=14),
        axis.title.y = element_text(size=14))
```

## **10 Different forecast methods - prediction from May 2023 to April 2025**

### **10.1 Mean forecast**

<h4>The Mean method predicts the future based on the mean from the dataset - in this case it's around 140 000 with expected variation within the confidence interval</h4>

```{r,fig.width=10,fig.height=6,fig.align='center'}
autoplot(meanf(df_ts,h=24)) + 
  scale_y_continuous(breaks = seq(round(min(df_ts),-4),round(max(df_ts),-4),by=20000)) +
  scale_x_continuous(breaks = seq(2012,2025,1))+
  theme_bw() + 
  th
```

```{r}
forecast(meanf(df_ts,h=24)) %>%
    kable(align = "c") %>%
    kable_styling(position = "center") %>%
    scroll_box(height = "100%")
```



### **10.2 Naive forecast** 

<h4> The naive method predicts the future based on the last value from the data set - in this case it's also around 140 000</h4>

```{r,fig.width=10,fig.height=6,fig.align='center'}
autoplot(naive(df_ts,h=24)) + 
  scale_x_continuous(breaks = seq(2012,2025,1))+
  theme_bw() +
  th
```

```{r}
forecast(naive(df_ts,h=24)) %>%
    kable(align = "c") %>%
    kable_styling(position = "center") %>%
    scroll_box(height = "100%")
```


### **10.3 Snaive forecast**

#### Snaive method predicts the future based on the last seasonal period (in this case there is a yearly seasonality). Therefore for 2 years prediction, the second year prediction is the same as for the first year**

```{r,fig.width=10,fig.height=6,fig.align='center'}
autoplot(snaive(df_ts,h = 24)) + 
  scale_x_continuous(breaks = seq(2012,2025,1))+
  theme_bw() + 
  th
```

```{r}
forecast(snaive(df_ts,h=24)) %>%
    kable(align = "c") %>%
    kable_styling(position = "center") %>%
    scroll_box(height = "100%")
```

### **10.4 Random walk with drift**

#### Random walk with drift is an extension of the trend that occures in the data set. This method implies, that the future data is based on the trend from the past

```{r,fig.width=10,fig.height=6,fig.align='center'}
autoplot(rwf(df_ts,h=24,drift=TRUE)) + 
  scale_x_continuous(breaks = seq(2012,2025,1))+
  theme_bw() + 
  th
```

```{r}
forecast(rwf(df_ts,h=24)) %>%
    kable(align = "c") %>%
    kable_styling(position = "center") %>%
    scroll_box(height = "100%")
```


### **10.5 ARIMA**

#### ARIMA is a model that:
<h4>1. Uses auto-regression and/or seasonal auto-regression (if needed)</h4>
<h4>2. Uses differencing to transform the time series from non-stationary to stationary (if needed)</h4>
<h4>3. Uses moving average (if needed), meaning that it uses residual lags</h4>


#### auto.arima function chooses the most optimal model, meaning a model with the lowest AIC score

```{r}
mod <- auto.arima(df_ts,trace = T)
```

```{r}
summary(mod)
```

#### ARIMA specification and important metrics

#### ARIMA(1,1,2)(0,0,2)[12] indicates:

<h4>1. First row of auto-regression</h4>
<h4>2. First row of differencing (to transform the data to stationary time series)</h4>
<h4>3. Second row of moving average</h4>
<h4>4. No seasonal auto-regression</h4>
<h4>5. No seasonal differencing</h4>
<h4>6. Second row of seasonal moving average</h4>
<h4>7. Seasonality occurs every 12 months</h4>

#### Important metrics:

<h4>1. MAE(mean absolute error)= indicates that the average absolute difference between actual value and its prediction = 10517.81</h4>
<h4>2. MAPE(mean absolute percentage error) indicates that the model was wrong by 7.61% on average</h4>
<h4>3. RMSE(root mean squared error) = indicates that the root of the average of squared differences between actual observation and its prediction = 13548.99</h4>

#### Formulas:

```{r,fig.width=5,fig.height=2,fig.align='center'}
ggplot() + 
    background_image(readPNG("images/MAE.png"))
```

```{r,fig.width=5,fig.height=2,fig.align='center'}
ggplot() + 
    background_image(readPNG("images/MAPE.png"))
```

```{r,fig.width=4,fig.height=1.5,fig.align='center'}
ggplot() + 
    background_image(readJPEG("images/rmse.jpg"))
```


#### ARIMA prediction (next 2 years: May 2023 - April 2025)

```{r,fig.width=10,fig.height=6,fig.align='center'}
autoplot(forecast(mod,h = 24))+
  scale_y_continuous(breaks = seq(round(min(df_ts),-4),round(max(df_ts),-4),by=20000))+
  scale_x_continuous(breaks = seq(2012,2025, by = 1))+
  theme_bw() + 
  th
```

```{r}
forecast(mod,h=24) %>%
    kable(align = "c") %>%
    kable_styling(position = "center") %>%
    scroll_box(height = "100%")
```

#### ARIMA residuals

#### The residuals appear to be randomly distributed. Ljung-box test results (p-value = 0.08817) indicates that the residuals are random, although 2 spikes from the ACF plot excess the insignifance area

```{r}
checkresiduals(mod)
```

#### Visualizing all prediction methods at once

```{r,fig.width=10,fig.height=6,fig.align='center'}
autoplot(df_ts) +
  autolayer(meanf(df_ts, h=24),
    series="Mean", PI=FALSE) +
  autolayer(naive(df_ts, h=24),
    series="Naïve", PI=FALSE) +
  autolayer(rwf(df_ts, drift=TRUE, h=24),
    series="Drift", PI=FALSE) +
  autolayer(snaive(df_ts,h = 24),
    series="Snaive", PI = FALSE) + 
  autolayer(forecast(mod,h = 24,level = FALSE),
    series="ARIMA",PI = FALSE) +
  ggtitle("Newly Registered Vehicles Prediction") +
  xlab("Year") + ylab("Newly registered vehicles") +
  guides(colour=guide_legend(title="Forecast")) + 
  scale_color_manual(values=c("#1BF31B","#FE5E02","#025AFE","#FB02FE","#FAFE02")) + 
  scale_x_continuous(breaks=seq(min(time(df_ts)),max(time(df_ts)+3)))+
  scale_y_continuous(breaks=seq(round(min(df_ts),-4),round(max(df_ts),-4),by=20000),labels=comma_format())+
  theme_bw()+
  theme(axis.text.x = element_text(size=14),
        axis.text.y = element_text(size=14),
        legend.title = element_text(size=18),
        legend.text = element_text(size=14),
        title = element_text(size=18),
        axis.title.x = element_text(size=14),
        axis.title.y = element_text(size=14))
```

## **9 Conclusions**

<h4>1. It is expected that the number of vehicles registered for the first time in Poland from May 2023 to April 2025 will remain relatively constant at around 140,000 per month</h4>
<h4>2. An increase to the 2018-2019 level is therefore unlikely, as is a decline to the values from around 2012</h4>
<h4>3. The most likely scenario is a a return to the 2016-2017 values</h4>

## **10 Sources**

1. [Vehicle classification] http://www.cepik.gov.pl/documents/76251/76577/Pojazdy+zarejestrowane+w+2022+r.+rodzajami+I-X+%28pdf%29/e35a777f-f2a1-472d-b3ae-9162f022543d
1. [2022 data] http://www.cepik.gov.pl/documents/76251/76577/Pojazdy+zarejestrowane+w+2022+r.+województwami+I-X+%28pdf%29/bc07beef-ae6c-41a7-acb3-49ff301ac6a8
1. [2021 data] http://www.cepik.gov.pl/documents/76251/76577/Pojazdy+zarejestrowane+w+2021+r.+województwami+I-XII+%28pdf%29/f04cf159-3611-4c61-b725-2d7fdf02f711
1. [2020 data] http://www.cepik.gov.pl/documents/76251/76577/Pojazdy+zarejestrowane+w+2020+r.+województwami+%28pdf%29+I-XII/12b09706-5453-4ce2-b941-2b00ff4ae8d2
1. [2019 data] http://www.cepik.gov.pl/documents/76251/76577/Pojazdy+zarejestrowane+w+2019+r.+województwami+%28pdf%29/4663a082-1e19-4cfd-aea1-f0d34e1f75b0
1. [2018 data] http://www.cepik.gov.pl/documents/76251/76577/Pojazdy+zarejestrowane+w+2018+r.+województwami+%28pdf%29/bcedb4a2-8256-47bd-bfbf-644e704d0fe2
1. [2017 data] http://www.cepik.gov.pl/documents/76251/76577/Pojazdy+zarejestrowane+w+2017+r.+województwami/ceb7ade4-db2f-43da-a8a3-a41683fe08bd
1. [2016 data] http://www.cepik.gov.pl/documents/76251/76577/Pojazdy+wg+województw+z+grudniem+2016/c5771bd3-4f24-4d90-8797-eed815d1fe8e
1. [2015 data] http://www.cepik.gov.pl/documents/76251/76577/Zarejestrowane+pojazdy+wg+województw+z+grudniem/7aae6743-c578-47ac-a2a8-8aabad6cdae7
1. [2014 data] http://www.cepik.gov.pl/documents/76251/76577/Pojazdy+zarejestrowane+2014/2c43af63-0f36-4c94-bc47-1605b7840796
1. [2013 data] http://www.cepik.gov.pl/documents/76251/76577/Pojazdy+zarejestrowane+w+2013+r/d66dcbf6-f2ce-4f94-96f6-7d9976c418e3
[2012 data] http://www.cepik.gov.pl/documents/76251/76577/Pojazdy+zarejestrowane+w+2012+r/9b58921a-8131-4830-a4c1-9b65955e1bde



