---
title: "CSGO Round T Side Win Prediction"
author: "Piotr Wieczorek"
date: "2023-10-23"
output: 
  html_document: 
    toc: yes
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
```

## Libraries 

```{r}
library(tidyverse)
library(randomForest)
library(janitor)
library(readxl)
library(caret)
library(glue)
library(hrbrthemes)
library(scales)
library(fastDummies)
library(coefplot)
library(gridExtra)
library(ROCit)
library(pROC)
```

```{r}
rm(list=ls())
```

## Reading and cleaning the data

```{r}
df <- read.csv('csgo_round_snapshots.csv')
```

### Looking for NA values

```{r}
sum(is.na(df)) #no NA's
```
### Checking colnames

```{r}
colnames(df)
```

```{r}
head(df)
```


```{r}
df %>%
  select_if(is.character) 
```

### Changing map and round_winner to factor, recoding bomb_planted to 0 and 1 

```{r}
df <- df %>%
  mutate(across(c("map","round_winner"),as.factor)) %>%
  mutate(bomb_planted = ifelse(bomb_planted == FALSE,0,1))
```

```{r}
head(df)
```

### Map frequency plot

```{r,fig.width=10,fig.height=6}
df %>%
  group_by(map) %>%
  summarise(map_freq = n()) %>%
  arrange(desc(map_freq)) %>%
  mutate(map=fct_reorder(map,map_freq)) %>%
  ggplot(aes(x=map,y=map_freq)) + 
  geom_bar(stat="identity") +
  geom_text(aes(label = map_freq), vjust = -0.5, size = 3.5) +
  labs(title = "Map Frequency") +
  theme_ft_rc()
```


### Round winner frequency plot 

```{r,fig.width=10,fig.height=6}
df %>%
  group_by(round_winner) %>%
  summarise(group_winner_freq = n()) %>%
  ggplot(aes(x=round_winner,y=group_winner_freq)) + 
  geom_bar(stat="identity") +
  geom_text(aes(label = group_winner_freq), vjust = -0.5, size = 3.5) +
  labs(title = "Round Winner Frequency") +
  ylab('Rounds won') + 
  xlab('Round winner')+
  theme_ft_rc()
```



### Analyzing T side weapons

```{r}
df %>%
  summarize(across(starts_with("t_weapon"),~ sum(.x)))
```

```{r}
df_t_weapons <- df %>%
  summarize(across(starts_with("t_weapon"),~ sum(.x))) %>%
  pivot_longer(everything()) %>%
  arrange(desc(value))
```

```{r}
df_t_weapons
```

### T pistol frequency plot

```{r,fig.width=10,fig.height=6}
df_t_weapons %>%
  filter(name %in% c("t_weapon_glock",
                     "t_weapon_deagle",
                     "t_weapon_tec9",
                     "t_weapon_p250")) %>%
  mutate(name = str_remove_all(string=name,pattern="t_weapon_")) %>%
  mutate(name = str_to_upper(name)) %>%
  mutate(name = fct_reorder(name,value)) %>%
  ggplot(aes(x=name,y=value)) +
  geom_bar(stat="identity") + 
  geom_text(aes(label = value), vjust = -0.5, size = 3.5) +
  labs(title = "T Pistol Weapons Frequency") +
  ylab('Pistol') + 
  xlab('Pistol frequency')+
  theme_ft_rc()
```


```{r}
table(df$bomb_planted) # there are only rounds in the data set where the bomb was planted
```

### T SMG weapons frequency

```{r,fig.width=10,fig.height=6}
df_t_weapons %>%
  filter(name %in% c("t_weapon_mac10",
                     "t_weapon_mp5sd",
                     "t_weapon_ump45",
                     "t_weapon_p90",
                     "t_weapon_bizon")) %>%
  mutate(name = str_remove_all(string=name,pattern="t_weapon_")) %>%
  mutate(name = str_to_upper(name)) %>%
  mutate(name = fct_reorder(name,value)) %>%
  filter(value >100) %>%
  ggplot(aes(x=name,y=value)) +
  geom_bar(stat="identity") + 
  geom_text(aes(label = value), vjust = -0.5, size = 3.5) +
  labs(title = "T SMG Weapons Frequency (only > 100)") +
  ylab('SMG') + 
  xlab('SMG frequency')+
  theme_ft_rc()
```

### T rifle weapons frequency

```{r,fig.width=10,fig.height=6}
df_t_weapons %>%
  filter(name %in% c("t_weapon_ak47","t_weapon_sg553","t_weapon_awp",
           "t_weapon_galilar","t_weapon_ssg08","t_weapon_g3sg1")) %>%
  mutate(name = str_remove_all(string=name,pattern="t_weapon_")) %>%
  mutate(name = str_to_upper(name)) %>%
  mutate(name = fct_reorder(name,value)) %>%
  filter(value > 100) %>%
  ggplot(aes(x=name,y=value)) +
  geom_bar(stat="identity") + 
  geom_text(aes(label = value), vjust = -0.5, size = 3.5) +
  labs(title = "T Rifle Weapons Frequency (only > 100)") +
  ylab('Rifle') + 
  xlab('Rifle frequency')+
  theme_ft_rc()
```


### Analyzing CT side weapons

```{r}
df_ct_weapons <- df %>%
  summarize(across(starts_with("ct_weapon"),~ sum(.x))) %>%
  pivot_longer(everything()) %>%
  arrange(desc(value))
```

```{r}
df_ct_weapons
```

### CT SMG weapons frequency

```{r,fig.width=10,fig.height=6}
df_ct_weapons %>%
  filter(name %in% c("ct_weapon_mp9","ct_weapon_ump45",
           "ct_weapon_p90","ct_weapon_mp5sd")) %>%
  mutate(name = str_remove_all(string=name,pattern="ct_weapon_")) %>%
  mutate(name = str_to_upper(name)) %>%
  mutate(name = fct_reorder(name,value)) %>%
  filter(value > 100) %>%
  ggplot(aes(x=name,y=value)) +
  geom_bar(stat="identity") + 
  geom_text(aes(label = value), vjust = -0.5, size = 3.5) +
  labs(title = "CT SMG Weapons Frequency (only > 100)") +
  ylab('SMG') + 
  xlab('SMG frequency')+
  theme_ft_rc()
```



### CT pistol weapons frequency

```{r,fig.width=10,fig.height=6}
df_ct_weapons %>%
  filter(name %in% c("ct_weapon_usps","ct_weapon_deagle","ct_weapon_p250",
           "ct_weapon_p2000","ct_weapon_c75auto","ct_weapon_fiveseven")) %>%
  mutate(name = str_remove_all(string=name,pattern="ct_weapon_")) %>%
  mutate(name = str_to_upper(name)) %>%
  mutate(name = fct_reorder(name,value)) %>%
  ggplot(aes(x=name,y=value)) +
  geom_bar(stat="identity") + 
  geom_text(aes(label = value), vjust = -0.5, size = 3.5) +
  labs(title = "CT Pistol Weapons Frequency") +
  ylab('Pistol') + 
  xlab('Pistol frequency')+
  theme_ft_rc()
```


### CT rifle weapons frequency

```{r,fig.width=10,fig.height=6}
df_ct_weapons %>%
  filter(name %in% c("ct_weapon_m4a4","ct_weapon_awp","ct_weapon_aug",
           "ct_weapon_m4a1s","ct_weapon_famas","ct_weapon_ssg08",
           "ct_weapon_scar20")) %>%
  mutate(name = str_remove_all(string=name,pattern="ct_weapon_")) %>%
  mutate(name = str_to_upper(name)) %>%
  mutate(name = fct_reorder(name,value)) %>%
  filter(value > 100) %>%
  ggplot(aes(x=name,y=value)) +
  geom_bar(stat="identity") + 
  geom_text(aes(label = value), vjust = -0.5, size = 3.5) +
  labs(title = "CT Rifle Weapons Frequency (only > 100)") +
  ylab('Rifle') + 
  xlab('Rifle frequency')+
  theme_ft_rc()
```

### Looking for near-zero-variance variables

```{r}
df %>%
select(12:97) %>%
  mutate(across(everything(),as.factor)) %>%
  summary()
```

```{r}
df%>%
  select(starts_with("ct_weapon")) %>%
  colnames()
```
### Deleting zero or near-zero-variance variables

```{r}
df <- df %>%
  select(-c("ct_weapon_bizon","t_weapon_bizon","t_weapon_elite",
            "ct_weapon_elite","ct_weapon_g3sg1","ct_weapon_m249",
            "t_weapon_m249","t_weapon_mag7","ct_weapon_mag7",
            "ct_weapon_mp5sd","t_weapon_mp5sd","ct_weapon_mp7",
            "t_weapon_mp7","t_weapon_mp9","ct_weapon_negev",
            "t_weapon_negev","ct_weapon_nova","t_weapon_nova",
            "ct_weapon_p90","t_weapon_p90","ct_weapon_r8revolver",
            "t_weapon_r8revolver", "ct_weapon_sawedoff","t_weapon_sawedoff",
            "ct_weapon_scar20","t_weapon_scar20","ct_weapon_xm1014",
            "t_weapon_xm1014","t_weapon_fiveseven","t_weapon_p2000",
            "ct_weapon_tec9","t_grenade_incendiarygrenade","ct_grenade_molotovgrenade",
            "ct_grenade_decoygrenade","t_grenade_decoygrenade"))
```

```{r}
df <- df %>%
  select(-c("bomb_planted"))
```

```{r}
head(df)
```


```{r}
df %>%
  select(11:61) %>%
  mutate(across(everything(), as.factor)) %>%
  summary()

### There are some issues with the data set, because it's impossible for the T side to have 6 players (there are 5 players in each team) and it's also extremely unlikely for the CT side to have 0 players alive (meaning all team members were dead) just once in 120 000 + observations.

### It would be pointless to include data about rounds where there are 0 CT players since it results in the end of the round and therefore the T side would win the round - we wouldn't have to consider other variables at all

### When there are 0 T players alive it wouldn't necesesserily (but most likely) mean that CT players have won the round because the T players can plant and detonate the bomb and die while doing so, but if the bomb explodes, T players win the round. 

### Because of the issues with the data set I decided to completely omit data where there are 0 alive players either from the CT side or the T side
```
### Filtering out data that doesn't make sense or has only single observation among 120 000 + observations

```{r}
df <- df %>%
  filter(!ct_players_alive %in% c(0)) %>%
  filter(!t_players_alive %in% c(0,6)) %>%
  filter(!t_weapon_cz75auto %in% c(6)) %>%
  filter(!t_grenade_flashbang %in% c(7)) %>%
  filter(!ct_grenade_smokegrenade %in% c(6)) %>%
  filter(!t_grenade_smokegrenade %in% c(9)) %>%
  filter(!ct_grenade_flashbang %in% c(6,7)) %>%
  filter(!ct_weapon_usps %in% c(6,7,8,9,10))
```

```{r}
unique(df$t_weapon_glock) # this also doesn't make sense - there are 5 players in the T side and they can have max 5 glocks total 
```

```{r}
df <- df %>% # these variables had almost no diversity or didn't make sense but I hadn't noticed that before
  select(-c("t_weapon_aug","t_weapon_famas","t_weapon_g3sg1",
            "ct_weapon_galilar","ct_weapon_glock","t_weapon_m4a1s",
            "ct_weapon_mac10","t_weapon_usps","ct_weapon_p2000"))
```

### Aggregating categories to further avoid near-zero-variance variables

```{r}
df <- df %>%
  mutate_if(is.numeric, as.factor)%>%
  mutate(ct_weapon_ak47 = fct_collapse(ct_weapon_ak47,
                                       "at_least_2" = c('2','3','4')),
         t_weapon_ak47 = fct_collapse(t_weapon_ak47,
                                      "at_least_4" = c('4','5')),
         ct_weapon_aug = fct_collapse(ct_weapon_aug,
                                      "at_least_2" = c('2','3','4','5')),
         ct_weapon_awp = fct_collapse(ct_weapon_awp,
                                      "at_least_2" = c('2','3','4','5')),
         t_weapon_awp = fct_collapse(t_weapon_awp,
                                     'at_least_1' = c('1','2','3','4')),
         ct_weapon_cz75auto = fct_collapse(ct_weapon_cz75auto,
                                           'at_least_2' = c('2','3','4','5')),
         t_weapon_cz75auto = fct_collapse(t_weapon_cz75auto,
                                          'at_least_2' = c('2','3','4','5')),
         ct_weapon_famas = fct_collapse(ct_weapon_famas,
                                        'at_least_2' = c('2','3','4','5')),
         t_weapon_galilar = fct_collapse(t_weapon_galilar,
                                       'at_least_2' = c('2','3','4','5')),
         ct_weapon_m4a1s = fct_collapse(ct_weapon_m4a1s,
                                        'at_least_1' = c('1','2','3')),
         ct_weapon_m4a4 = fct_collapse(ct_weapon_m4a4,
                                       'at_least_3' = c('3','4','5')),
         t_weapon_m4a4 = fct_collapse(t_weapon_m4a4,
                                      'at_least_1' = c('1','2','3')),
         t_weapon_mac10  = fct_collapse(t_weapon_mac10,
                                        'at_least_2' = c('2','3','4','5')),
         ct_weapon_mp9 = fct_collapse(ct_weapon_mp9,
                                      'at_least_2' = c('2','3','4','5')),
         ct_weapon_sg553 = fct_collapse(ct_weapon_sg553,
                                        'at_least_2' = c('2','3','4')),
         t_weapon_sg553 = fct_collapse(t_weapon_sg553,
                                       'at_least_2' = c('2','3','4','5')),
         ct_weapon_ssg08 = fct_collapse(ct_weapon_ssg08,
                                        'at_least_1' = c('1','2','3','4')),
         t_weapon_ssg08 = fct_collapse(t_weapon_ssg08,
                                       'at_least_1' = c('1','2','3')),
         ct_weapon_ump45 = fct_collapse(ct_weapon_ump45,
                                        'at_least_1' = c('1','2','3')),
         t_weapon_ump45 = fct_collapse(t_weapon_ump45,
                                       'at_least_1' = c('1','2','3')),
         ct_weapon_deagle = fct_collapse(ct_weapon_deagle,
                                         'at_least_3' = c('3','4','5')),
         t_weapon_deagle = fct_collapse(t_weapon_deagle,
                                        'at_least_3' = c('3','4','5')),
         ct_weapon_fiveseven = fct_collapse(ct_weapon_fiveseven,
                                            'at_least_2' = c('2','3','4','5')),
         ct_weapon_p250 = fct_collapse(ct_weapon_p250,
                                       'at_least_2' = c('2','3','4','5')),
         t_weapon_p250  = fct_collapse(t_weapon_p250,
                                       'at_least_2' = c('2','3','4','5')),
         t_weapon_tec9 = fct_collapse(t_weapon_tec9,
                                      'at_least_1' = c('1','2','3','4')),
         t_grenade_hegrenade = fct_collapse(t_grenade_hegrenade,
                                            'at_least_3' = c('3','4','5')))
```

```{r}
df <- df %>%
  mutate(across(c("ct_score","t_score","ct_health",
                  "t_health","ct_armor","t_armor",
                  "ct_money","t_money","time_left"),as.numeric))
```

## Modeling part

### Data partition

```{r}
set.seed(778)
data_partition <- createDataPartition(df$round_winner, p = 0.75, list=FALSE)

train_set <- df[data_partition,]
test_set <- df[-data_partition,]

print(nrow(train_set))
print(nrow(test_set))
```

```{r}
head(train_set)
```

```{r}
head(test_set)
```
### Creating dummy variables

```{r}
train_set_w_dummies <- fastDummies::dummy_cols(train_set[,-52], remove_first_dummy = TRUE, remove_selected_columns = TRUE,omit_colname_prefix = TRUE)

train_set <- cbind(train_set_w_dummies,train_set$round_winner)

test_set_w_dummies <- fastDummies::dummy_cols(test_set[,-52], remove_first_dummy = TRUE, remove_selected_columns = TRUE,omit_colname_prefix = TRUE)

test_set <- cbind(test_set_w_dummies,test_set$round_winner)
```


```{r}
head(train_set)
```


```{r}
colnames(train_set)[139] <- 'round_winner'
colnames(test_set)[139] <- 'round_winner'
```

### Model formula


```{r}
first_model <- glm(formula = round_winner ~ ., data = train_set, family = "binomial")
summary(first_model)
```

### Prediction

```{r}
prediction <- predict(object = first_model, test_set, type = "response")
```


```{r}
head(prediction)
```


```{r}
prediction_group <- ifelse(prediction >= 0.5, "T", "CT")
```

```{r}
head(prediction_group)
```


### Confusion matrix


```{r}
confusionMatrix(data = as.factor(prediction_group), reference = test_set$round_winner,
    positive = "T")
```

### exponent of coefficients

```{r}
exp(first_model$coefficients)
```

### ROC curve

```{r}
sim_roc <- roc(response = test_set$round_winner, predictor = prediction,
    levels = c("T", "CT"))
```

```{r}
ggroc(sim_roc, legacy.axes = TRUE) + annotate("text", x = 0.5,
    y = 0.5, label = paste0("AUC: ", round(auc(sim_roc), digits = 2))) +
    theme_minimal()
```

