---
title: "Palmer Penguins - Comparing Body Weight Medians with Regard to Islands"
author: "Piotr Wieczorek"
date: "2023-12-08"
output: 
  html_document: 
    toc: yes
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```


```{r}
rm(list=ls())
```


## Libraries

```{r}
library(tidyverse)
library(naniar)
library(scales)
library(palmerpenguins)
library(missForest)
library(skimr)
library(car)
library(rstatix)
library(ggpubr)
library(png)
```



## Loading and cleaning the data


```{r}
df <- palmerpenguins::penguins
```


### Missing data imputation


```{r}
naniar::miss_var_summary(df) %>%
  filter(n_miss > 0)
```

```{r}
naniar::gg_miss_var(df[,!(colnames(df) %in% c("year","species","island"))])
```

```{r}
df <- missForest(as.data.frame(df))
```


```{r}
df$OOBerror
```

```{r}
df <- df$ximp
```



### Mutating columns


```{r}
df <- df %>%
  mutate_if(is.numeric,~round(.,2)) %>%
  mutate(sex = as.character(sex)) %>%
  mutate(sex = str_to_sentence(sex)) %>%
  mutate(sex = as.factor(sex))
```


## Exploratory analysis


### Exploring body_mass_g by island

```{r}
th <- theme(plot.title = element_text(size=18),
            axis.title.x = element_text(size=16),
            axis.title.y=element_text(size=16),
            axis.text.x= element_text(size=14),
            axis.text.y=element_text(size=14),
            strip.text = element_text(size=16),
            legend.title =  element_text(size=18),
            legend.text  = element_text(size=14))
```


```{r,fig.width=12,fig.heigth=10}
df %>%
  select(c(island,body_mass_g,sex)) %>%
  filter(sex=="Male") %>%
  ggplot(aes(x=island,y=body_mass_g,fill=island,color=island)) + 
  geom_boxplot()  +
  geom_jitter(shape=21,size=3,
              position = position_jitterdodge(dodge.width = 0.75, jitter.width = 0.5))+ 
  scale_fill_manual(values=c("#EBF1AD","#ADF1CB","#70E1E7")) + 
  scale_color_manual(values=c("#C8CD8E","#80B18B","#2D9BA0")) +
  labs(title = "Body Mass Boxplots by Island",x="",y="")+
  theme_bw() + th
```



### Exploring body_mass_g distributions by island

 
```{r}
df2<-df %>%
  select(c(island,body_mass_g,sex)) %>%
  filter(sex=="Male") %>%
  select(c(1:2))
```

```{r}
head(df2)
```


```{r,fig.width=14,fig.heigth=10}
df2 %>%
  ggplot(aes(x=body_mass_g,fill=island,color=island)) + 
  geom_density() + 
  scale_fill_manual(values=c("#EBF1AD","#ADF1CB","#70E1E7")) + 
  scale_color_manual(values=c("#C8CD8E","#80B18B","#2D9BA0")) +
  facet_wrap(~island,scales="free") + 
  labs(title = "Body Mass Density Plots by Island",x="",y="")+
  theme_bw() + th
```


## Checking statistical assumptions for ANOVA test - normal distributions and homoscedasticity


### Normality assumption


```{r}
shapiro.test(df2[df2$island=="Biscoe","body_mass_g"])
```

```{r}
shapiro.test(df2[df2$island=="Dream","body_mass_g"])
```

```{r}
shapiro.test(df2[df2$island=="Torgersen","body_mass_g"])
```

### Conclusion

#### body_mass_g is not normally distributed among penguins living on the Biscoe island


```{r}
leveneTest(body_mass_g ~ island, data= df2)
```


### Conclusion

#### The variance is not equal among islands


## Kruskal-Wallis test


### Formula
```{r,fig.width=8,fig.height=3}
ggplot() + 
    background_image(readPNG("images/kwtest.png"))
```



```{r}
kruskal_wallis_test <- kruskal.test(formula = body_mass_g ~ island, data = df2)
```



```{r}
kruskal_wallis_test
```


### Conclusion

#### The medians among penguins living on different islands are not equal at least for one pair of islands


## Dunn test
```{r}
dunn_test(data=df2,formula=body_mass_g~island)
```


### Conclusion

#### 1. The body_mass_g medians among penguins living on the Dream and Torgersen island are not different in terms of statistical signifance
#### 2. The body_mass_g medians among penguins living on the Dream and Biscoe island are different in terms of statistical significane
#### 3. The body_mass_g medians among penguins living on the Biscoe and Torgersen island are different in terms of statistical signifance
