---
title: "Monte Carlo Simulation - 11 Bit Return Rate Prediction"
author: "Piotr Wieczorek"
date: "2023-10-07"
output: 
  html_document: 
    toc: yes
    theme: readable
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message = FALSE)
```

## Loading libraries

```{r}
library(tidyverse)
library(lubridate)
library(tseries)
library(PerformanceAnalytics)
library(quantmod)
```

## Loading and cleaning the data about 11 Bit daily return rates

```{r}
rm(list=ls())
```

```{r}
df <- read.csv('11b_d.csv')
```

```{r}
head(df)
```

```{r}
df <- df %>%
  select(c("Data","Zamkniecie")) %>%
  mutate(Data = lubridate::ymd(Data))
```

```{r}
colnames(df)
```
```{r}
colnames(df) <- c("Date","Close")
```

```{r}
head(df,10)
```


```{r}
data_xts <- xts(df[,-1],order.by = df$Date)
```


```{r}
head(data_xts,10)
```
```{r}
colnames(data_xts) <- c("Close")
```



```{r}
data_xts$log_return_rate <- PerformanceAnalytics::Return.calculate(data_xts$Close,method="log")
```


```{r}
head(data_xts,10)
```

## Exploring the data

### 11 Bit stock rate plot

```{r}
chartSeries(data_xts$Close,name = '11 Bit Daily Stock Price')
```

```{r}
chartSeries(data_xts$log_return_rate, name = '11 Bit Daily Logarithmic Return Rate')
```


### 11 Bit daily log return rate density plot

```{r}
data_xts[-1,] %>%
  ggplot(aes(x=log_return_rate)) + 
  geom_density() +
  labs(title="11 Bit Logarithmic Daily Return Rate Density") + 
  theme_dark() 
```

### QQ Plot for the log daily return rate

```{r}
data_xts %>%
  ggplot(aes(sample = log_return_rate)) +
  stat_qq()  +
  stat_qq_line()
```

### Stationarity test

```{r}
adf.test(data_xts[-1,]$log_return_rate)
```

#### 11 Bit daily return rate is of stationary nature

## Daily Return Rate Monte Carlo Simulation

```{r}
X = matrix(ncol = 30,nrow = 10000)
for(day in 1:30) {
  # for each day (30 in total) sample the return rate 100000 times
  for(i in 1:10000){
    X[i,day] = sample(data_xts[-1,]$log_return_rate,1)
  }
}
```

```{r}
daily_log_return_rates <- apply(X,2,mean)
```

```{r}
daily_log_return_rates
```
### Return Rate for 30 (traiding) day period

```{r}
sum(daily_log_return_rates)
```
### Annual Return Rate

```{r}
PerformanceAnalytics::Return.annualized(R = daily_log_return_rates,scale=252)
```
#### This result is almost identical to manual calculations

```{r}
0.03413412/30
```

```{r}
(1+0.001137804)^252-1
```
#### And it's identical if we pass the average daily return rate to the formula

```{r}
PerformanceAnalytics::Return.annualized(R=0.001137804,scale=252)
```

