---
title: "Penguins Classification"
author: "Piotr Wieczorek"
date: "2023-11-26"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

```{r}
rm(list=ls())
```

## Libraries
```{r}
library(tidyverse)
library(caret)
library(naniar)
library(janitor)
library(palmerpenguins)
library(missForest)
library(skimr)
library(gridExtra)
library(randomForest)
library(fastDummies)
```

## Loading and cleaning the data set

```{r}
df <- palmerpenguins::penguins
```

```{r}
head(df)
```

```{r}
glimpse(df)
```

### NA values summary

```{r}
naniar::miss_var_summary(df)
```

```{r}
naniar::gg_miss_var(df)
```

### Data imputation


```{r}
df <- missForest::missForest(as.data.frame(df))
```

#### Out of bag error rate

```{r}
df$OOBerror
```

```{r}
df <- df$ximp
```


```{r}
sum(is.na(df))
```

```{r}
head(df)
```

### Rounding numerical data

```{r}
df <- df %>%
  mutate_if(is.numeric,~round(.,2))
```

```{r}
head(df)
```


### Mutating the data
```{r}
df <- df %>%
  mutate(across(where(is.factor),as.character)) %>%
  mutate(across(where(is.character),~str_to_sentence(.))) %>%
  mutate(across(where(is.character),as.factor)) %>%
  mutate(across(where(is.numeric),~round(.,1))) %>%
  mutate(sex = fct_recode(sex, M = "Male", `F` = "Female")) %>%
  mutate(year = as.factor(year))
```


## Exploratory analysis

### Numerical data summary

```{r}
df %>%
  select_if(is.numeric) %>%
  skim()
```


#### Means of numerical data by island

```{r}
df_island <- df %>%
  group_by(island) %>%
  summarize(across(where(is.numeric),~mean(.),.names="{col}_mean")) %>%
  mutate(across(where(is.numeric),~round(.,2)))
```

#### Means of numerical data by species

```{r}
df_island <- df %>%
  group_by(species) %>%
  summarize(across(where(is.numeric),~mean(.),.names="{col}_mean")) %>%
  mutate(across(where(is.numeric),~round(.,2)))
```

### Factor variables summary

```{r}
df %>%
  select_if(is.factor) %>%
  summary()
```


### Data visualization

#### Boxplots of numerical data by species

```{r,fig.width=10,fig.height=6}
th <- theme(axis.text.x = element_text(size=18),
            axis.text.y = element_text(size=18),
            axis.title.x = element_text(size=20),
            axis.title.y = element_text(size=20),
            title = element_text(size=22),
            strip.text = element_text(size=24),
            legend.text = element_text(size=18),
            legend.title = element_text(size=22)
            )

df %>%
  select(c("species","bill_length_mm","bill_depth_mm","flipper_length_mm","body_mass_g")) %>%
  gather(c(2:5),value="val",key="key") %>%
  ggplot(aes(x=species, y = val, fill = species)) + 
  geom_boxplot(outlier.shape=NA) + 
  geom_jitter(aes(color=species),alpha=0.35) + 
  facet_wrap(~key, scales="free") +theme_light() + theme(strip.text = element_text(size=16))
```

#### Boxplots of numerical data by species, island and sex

```{r,fig.width=20,fig.height=14}

grid.arrange(
  
  df %>%
  ggplot(aes(x=species,y=bill_length_mm,fill=sex)) +
  geom_boxplot() + 
    facet_wrap(~island, scales="free") +theme_light() + th,
  
  df %>%
    ggplot(aes(x=species,y=bill_depth_mm,fill=sex)) +
    geom_boxplot() + 
    facet_wrap(~island, scales="free") + theme_light() + th,
  
  df %>%
    ggplot(aes(x=species, y = flipper_length_mm,fill=sex)) + 
    geom_boxplot() + 
    facet_wrap(~island, scales="free")+ theme_light() + th,
  
  df %>%
    ggplot(aes(x=species,y = body_mass_g, fill=sex)) + 
    geom_boxplot() + 
    facet_wrap(~island, scales="free") + theme_light() + th ,
  
  ncol = 2, nrow = 2
  
)
```


### Density plots by species and sex
```{r,fig.width=12,fig.height=6}
df %>%
  select(c("species","bill_length_mm","bill_depth_mm","flipper_length_mm","body_mass_g","sex")) %>%
  gather(-c("species","sex"),value="val",key="key") %>% 
  ggplot(aes(x=val,fill=species))  +
  geom_density(alpha=0.35) + 
  facet_wrap(~key + sex,scales="free",nrow = 2,ncol=4) + theme_light() + theme(strip.text = element_text(size=16))
```



## Modeling

### Data partition


```{r}
data_partition <- caret::createDataPartition(df$species, p = 0.7, list=FALSE)

train_set <- df[data_partition,]
test_set <- df[-data_partition,]

print(dim(train_set))
print(dim(test_set))
```

### Creating dummy variables

```{r}
train_dummies <-fastDummies::dummy_cols(train_set[,2:8], remove_first_dummy = TRUE, remove_selected_columns = TRUE)
train_set <- cbind(train_dummies,train_set[,1])

test_dummies <- fastDummies::dummy_cols(test_set[,2:8],remove_first_dummy = TRUE, remove_selected_columns = TRUE)
test_set <- cbind(test_dummies, test_set[,1])
```


```{r}
colnames(train_set)[10] <- "species"
colnames(test_set)[10] <- "species"
```

```{r}
head(train_set)
```

```{r}
head(test_set)
```

### Random forest model

```{r}
rf_mod <- randomForest(species ~. -year_2008 -year_2009,data = train_set, ntree=20, mtry = floor(sqrt(ncol(train_set))),importance = TRUE)
rf_mod
```

#### Error plot


```{r}
plot(rf_mod)
```

#### Importance plot

```{r}
varImpPlot(rf_mod,type=1)
```
#### Prediction

```{r}
head(predict(rf_mod,test_set))
```

#### Confusion matrix based on the predictions

```{r}
confusionMatrix(predict(rf_mod,test_set),test_set$species)
```


