---
title: "Predicting Diamonds Price - Linear Regression Example"
author: "Piotr Wieczorek"
date: "2023-10-20"
output: 
  html_document: 
    toc: yes
    highlight: monochrome
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
```

```{r}
rm(list=ls())
```


## Libraries 

```{r}
library(tidyverse)
library(scales)
library(glue)
library(skimr)
library(tableone)
library(janitor)
library(caret)
library(fastDummies)
```

## Reading and exploring the data

```{r}
df <- diamonds
```

### Overall summary

```{r}
summary(df)
```
### Looking for NA values

```{r}
colSums(is.na(df))
```

```{r}
factor_colnames <- df %>%
  select_if(is.factor) %>%
  colnames()
```

```{r}
factor_colnames
```
### Summary of factor variables

```{r}
tableone::CreateCatTable(data=df,vars=factor_colnames)
```

### Aggregating least frequent categories

```{r}
df <- df %>%
  mutate(clarity = fct_collapse(clarity,
                                "Other_clarity" = c("I1","IF")),
         cut = fct_collapse(cut,
                            "Other_cut" = c("Fair","Good")))
```

### All numerical variable's density plots

```{r,fig.width=10,fig.height=6}
df %>%
  select(c(1,5:10)) %>%
  gather(1:7,key="key",value="val") %>%
  ggplot(aes(x=val)) +
  geom_density() +
  facet_wrap(~key,scales="free") + 
  theme_light()
```

### Taking the log out of price

```{r}
df[,c("price")] <- sapply(df[,c("price")], function(x) log(x))
```


```{r}
colnames(df)[7] <- "price_log"
```


```{r}
df %>%
  select_if(is.numeric) %>%
  head()
```

### Exploring correlation between numerical variables

```{r}
cor(df%>%select_if(is.numeric),method = "spearman")
```
### Deleting variables based on the correlation matrix

```{r}
df<-df%>%
  select(-c("y","z"))
```


```{r}
cor(df%>%select_if(is.numeric))
```
### Deleting outliers from the x variable

```{r}
x_outliers <- boxplot.stats(df$x)$out
x_outliers_ind <- which(df$x %in% c(x_outliers))
print(length(x_outliers)) # number of x outliers
```
```{r}
df <- df[-x_outliers_ind,] %>%
  select(-c("table","depth","carat"))
```

### Boxplot of x after removing 32 outliers

```{r}
df %>%
  ggplot(aes(x=x)) + 
  geom_boxplot()
```

### Boxplot of price after taking the log out of it

```{r}
df %>%
  ggplot(aes(x=price_log)) +
  geom_boxplot()
```

```{r}
df$cut <- factor(df$cut, ordered = FALSE)
df$color <- factor(df$color,ordered = FALSE)
df$clarity <- factor(df$clarity,ordered=FALSE)
```

## Modelling

### Data partition 

```{r}
set.seed(123)
data_partition <- caret::createDataPartition(df$price_log,p=0.7,list=FALSE)

train_set <- df[data_partition,]
test_set <- df[-data_partition,]
```



```{r}
print(dim(train_set))
print(dim(test_set))
print(nrow(train_set)/(nrow(train_set)+nrow(test_set)))
```
### Relationship between x and price_log

```{r}
df %>%
  ggplot(aes(x=x,y=price_log)) +
  geom_point() +
  labs(title='X vs Price_log')
```

### Creating train set

```{r}
train_set <- dummy_cols(train_set,remove_first_dummy = TRUE,remove_selected_columns = TRUE)
```


### Creating test set


```{r}
test_set <- dummy_cols(test_set,remove_first_dummy = TRUE,remove_selected_columns = TRUE)
```

### Model formula

```{r}
mod <- lm(formula = price_log ~.,data=train_set)
print(summary(mod))
```

```{r}
(exp(coef(mod)["x"]) - 1) * 100 # for every 1 increase in X, price increases by 161%
```
```{r}
exp(coef(mod))
```
### Interpretation
1. for every 1 increase in X, price increases by 161%
2. On average, Very Good cut diamonds prices are 1.037 times the reference cut diamonds prices
3. On average, Premium cut diamonds prices are 0.999 times the reference cut diamonds prices
4. On average, Ideal cut diamonds prices are 1.0608 times the reference cut diamonds prices
5. On average, color E diamonds prices are 0.9418 times the reference color diamonds prices
6. On average, color F diamonds prices are 0.9217 times the reference color diamonds prices
7. On average, color G diamonds prices are 0.8695 times the reference color diamonds prices
8. On average, color H diamonds prices are 0.7725 times the reference color diamonds prices
9. On average, color I diamonds prices are 0.6799 times the reference color diamonds prices
10. On average, color J diamonds prices are  0.5910 times the reference color diamonds prices
11. On average, clarity SI2 diamonds prices are 0.7336 times the reference clarity diamonds prices
12. On average, clarity SI1 diamonds prices are 0.8756 times the reference clarity diamonds prices
13. On average, clarity VS2 diamonds prices are 0.9930 times the reference clarity diamonds prices
14. On average, clarity VS1 diamonds prices are 1.058 times the reference clarity diamonds prices
15. On average, clarity VVS2 diamonds prices are 1.174 times the reference clarity diamonds prices
16. On average, clarity VVS1 diamonds prices are 1.2362 times the reference clarity diamonds prices


### Model diagnostics

```{r,fig.width=10,fig.height=6}
par(mfrow=c(2,2))
plot(mod)
```





